### 2024-01-04 manuscript.Rmd
test commit

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: html_document
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Abstract
========
Background
----------
Trauma is a leading cause of mortality and morbidity worldwide. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

Methods *redigera*
-------
This was a registry-based study using all patients between 2014 and 2022 from the Karoliniska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.

Results
-------

Conclusion
----------

Introduction
============
Trauma is a leading cause of mortality and morbidity worldwide being the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly [1]. A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions. Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2]. Examples of opportunities for improvement include lack of resources and management errors.

Albaaj et al. recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna. The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [3]. Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.

According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury. Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [4]. These are examples of factors that could be important in care on a higher hospital level such as the ICU. The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.

```{r}

@ARTICLE{1,
  title     = "Short-term versus long-term trauma mortality: A systematic
               review",
  author    = "Frydrych, Lynn M and Keeney-Bonthrone, Toby P and Gwinn,
               Elizabeth and Wakam, Glenn K and Anderson, Maia S and Delano,
               Matthew J",
  journal   = "J. Trauma Acute Care Surg.",
  publisher = "Ovid Technologies (Wolters Kluwer Health)",
  volume    =  87,
  number    =  4,
  pages     = "990--997",
  month     =  oct,
  year      =  2019,
  language  = "en"
}

#https://pubmed.ncbi.nlm.nih.gov/31589196/
#Article on the evolution of long-term mortality in trauma patients as compared to the improved mortality numbers short-term

@ARTICLE{2,
  title    = "Identification of preventable trauma deaths: confounded
              inquiries?",
  author   = "Wilson, D S and McElligott, J and Fielding, L P",
  abstract = "The published evaluation of methods for identifying preventable
              trauma deaths contains many unstudied confounding factors. To
              investigate the reliability of methods for identifying such
              preventable deaths, we compared three consensus systems using
              separate five-member general review panels assessing 20
              non-central nervous system fatalities: panel A, independent
              judgments; panel B, discussion of all cases preceding individual
              judgments; and panel C, independent judgments followed by
              discussion and equivocal case reassignment. The Kappa concordance
              index was low for all methods (method A, 0.20; methods B and C,
              0.40). Of the 11 deaths judged preventable by at least one panel,
              only one death was judged preventable by all three panels.
              Consensus agreement (four of five assessors) was 20\% for panel
              A, 45\% for panel B, and 10\% for panel C (difference between
              panels B and C, p less than 0.03). In panel C, discussion
              affected the rate of equivocal case designation from 30\% to 5\%.
              Thus different consensus methods yielded different results. We
              conclude that individual case review can be severely flawed and
              therefore should not be used to measure institutional quality of
              patient care. We recommend that assessment of institutional
              performance should be based on objective evaluation methods,
              which require the study of patient population outcomes, rather
              than on subjective methods in which individual cases are
              reviewed.",
  journal  = "J. Trauma",
  volume   =  32,
  number   =  1,
  pages    = "45--51",
  month    =  jan,
  year     =  1992,
  language = "en"
}

#https://journals.lww.com/jtrauma/abstract/1992/01000/identification_of_preventable_trauma_deaths_.10.aspx

@ARTICLE{3,
  title    = "Patient and process factors associated with opportunities for
              improvement in trauma care: a registry-based study",
  author   = "Albaaj, Hussein and Attergrim, Jonatan and Str{\"o}mmer, Lovisa
              and Brattstr{\"o}m, Olof and Jacobsson, Martin and Wihlke,
              Gunilla and V{\"a}sterbo, Liselott and Joneborg, Elias and Gerdin
              W{\"a}rnberg, Martin",
  abstract = "BACKGROUND: Trauma is one of the leading causes of morbidity and
              mortality worldwide. Morbidity and mortality review of selected
              patient cases is used to improve the quality of trauma care by
              identifying opportunities for improvement (OFI). The aim of this
              study was to assess how patient and process factors are
              associated with OFI in trauma care. METHODS: We conducted a
              registry-based study using all patients between 2017 and 2021
              from the Karolinska University Hospital who had been reviewed
              regarding the presence of OFI as defined by a morbidity and
              mortality conference. We used bi- and multivariable logistic
              regression to assess the associations between the following
              patient and process factors and OFI: age, sex, respiratory rate,
              systolic blood pressure, Glasgow Coma Scale (GCS), Injury
              Severity Score (ISS), survival at 30 days, highest hospital care
              level, arrival on working hours, arrival on weekends, intubation
              status and time to first computed tomography (CT). RESULTS: OFI
              was identified in 300 (5.8\%) out of 5182 patients. Age, missing
              Glasgow Coma Scale, time to first CT, highest hospital care level
              and ISS were statistically significantly associated with OFI.
              CONCLUSION: Several patient and process factors were found to be
              associated with OFI, indicating that patients with moderate to
              severe trauma and those with delays to first CT are at the
              highest odds of OFI.",
  journal  = "Scand. J. Trauma Resusc. Emerg. Med.",
  volume   =  31,
  number   =  1,
  pages    = "87",
  month    =  nov,
  year     =  2023,
  keywords = "Acute surgery; Opportunities for improvement; Trauma; Trauma care",
  language = "en"
}

#Tidigare publicerad artikel av NOACs
#Albaaj, H. et al. Patient and process factors associated with opportunities for improvement in trauma care: a registry based study. Scandinavian Journal of Trauma, Resuscitation and Emergency Medicine (2023) 31:87.https://doi.org/10.1186/s13049-023-01157-y.

@ARTICLE{4,
  title    = "National assessment of opportunities for improvement in
              preventable trauma deaths: A mixed-methods study",
  author   = "Kwon, Junsik and Lee, Myeonggyun and Jung, Kyoungwon",
  abstract = "Trauma is a significant public health issue worldwide,
              particularly affecting economically active age groups. Quality
              management of trauma care at the national level is crucial to
              improve outcomes of major trauma. In Korea, a biennial nationwide
              survey on preventable trauma death rate is conducted. Based on
              the survey results, we analyzed opportunities for improving the
              trauma treatment process. Expert panels reviewed records of 8282
              and 8482 trauma-related deaths in 2017 and 2019, respectively,
              identifying 258 and 160 cases in each year as preventable deaths.
              Opportunities for improvement were categorized into prehospital,
              interhospital, and hospital stages. Hemorrhage was the primary
              cause of death, followed by sepsis/multiorgan failure and central
              nervous system injury. Delayed hemostatic procedures and
              transfusions were common areas for improvement in hospital stage.
              Interhospital transfers experienced significant delays in arrival
              time. This study emphasizes the need to enhance trauma care by
              refining treatment techniques, centralizing patients in
              specialized facilities, and implementing comprehensive reviews
              and performance improvements throughout the patient transfer
              system. The findings offer valuable insights for addressing
              trauma care improvement from both clinical and systemic
              perspectives.",
  journal  = "Healthcare (Basel)",
  volume   =  11,
  number   =  16,
  month    =  aug,
  year     =  2023,
  keywords = "mortality; patient transfer; quality improvement; treatment
              outcome; wounds and injuries",
  language = "en"
}

#https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10454732/pdf/healthcare-11-02291.pdf

```

Methods 
=======
Study design
------------
We conducted single-centre registry-based retrospective cohort study using data from the trauma registry and trauma care quality database at Karolinska University Hospital in Solna. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome of the selected patients was OFI identified by the multidisciplinary review board and then recorded in the trauma care quality database. From 2017 the trauma care quality database includes all patients in the trauma registry as well as the results of the mortality and morbidity review. This project will use these two databases and assess what types of OFI occurred in ICU patients and how different patient factors; such as age, sex, mechanism of injury and injury severity, are associated with OFI using logistic regression. A 5% significance level and 95% confidence levels will be used.

Setting
-------
The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there. The hospital has direct access to radiology, surgery, intensive care and consultants from relevant specialities [6]. All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses. The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI include the need for better organization or more senior members taking part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.

```{r}

@ARTICLE{5,
  title    = "Trauma center levels defined",
  author   = "Trauma Center Association of America",
  URL  = "https://www.traumacenters.org/page/TraumaCenterLevels#Goals",
  year     =  2014,
#Hur källhänvisar man webbsida?

@ARTICLE{6,
  title    = "Traumamanual",
  author   = "Karolinska Universitetssjukhuset Solna",
  year     =  2020,
#Include ISS? And audit filters?

```

Participants
------------
The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9. We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022*. *We excluded patients who were younger than 15 years and patients who were dead on arrival.*

Variables and data sources/measurements
---------------------------------------
Study outcome
The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors. *vilka OFIs ska läggas till för att analyseras?*

Patient and process factors 
The factors were chosen from the trauma registry, based on the locally used audit filters, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry. The categorical factors were *sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated*. The continuous factors age and Injury Severity Score (ISS) were also included. Working hours was defined as between 8.00 a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.

Statistical analysis 
Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level. The programming language used was R for all analyses []. The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.

*lägg till?*
A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above. We present sample characteristics using descriptive statistics. Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.

```{r}
# LÄNKA till http://www.rstudio.com/ 
```

Bias 
----
*senare*

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------
*senare*

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------
*senare*

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------
*senare*

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d) Cohort study: If applicable, explain how loss to follow-up was addressed

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2024-01-05 manuscript.Rmd
updated 5/1

@@
 
-Methods *redigera*
+Methods
 -------
 This was a registry-based study using all patients between 2014 and 2022 from the Karoliniska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.
+Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...
 
@@
 
-```{r}
-
-@ARTICLE{1,
-  title     = "Short-term versus long-term trauma mortality: A systematic
-               review",
-  author    = "Frydrych, Lynn M and Keeney-Bonthrone, Toby P and Gwinn,
-               Elizabeth and Wakam, Glenn K and Anderson, Maia S and Delano,
-               Matthew J",
-  journal   = "J. Trauma Acute Care Surg.",
-  publisher = "Ovid Technologies (Wolters Kluwer Health)",
-  volume    =  87,
-  number    =  4,
-  pages     = "990--997",
-  month     =  oct,
-  year      =  2019,
-  language  = "en"
-}
-
-#https://pubmed.ncbi.nlm.nih.gov/31589196/
-#Article on the evolution of long-term mortality in trauma patients as compared to the improved mortality numbers short-term
-
-@ARTICLE{2,
-  title    = "Identification of preventable trauma deaths: confounded
-              inquiries?",
-  author   = "Wilson, D S and McElligott, J and Fielding, L P",
-  abstract = "The published evaluation of methods for identifying preventable
-              trauma deaths contains many unstudied confounding factors. To
-              investigate the reliability of methods for identifying such
-              preventable deaths, we compared three consensus systems using
-              separate five-member general review panels assessing 20
-              non-central nervous system fatalities: panel A, independent
-              judgments; panel B, discussion of all cases preceding individual
-              judgments; and panel C, independent judgments followed by
-              discussion and equivocal case reassignment. The Kappa concordance
-              index was low for all methods (method A, 0.20; methods B and C,
-              0.40). Of the 11 deaths judged preventable by at least one panel,
-              only one death was judged preventable by all three panels.
-              Consensus agreement (four of five assessors) was 20\% for panel
-              A, 45\% for panel B, and 10\% for panel C (difference between
-              panels B and C, p less than 0.03). In panel C, discussion
-              affected the rate of equivocal case designation from 30\% to 5\%.
-              Thus different consensus methods yielded different results. We
-              conclude that individual case review can be severely flawed and
-              therefore should not be used to measure institutional quality of
-              patient care. We recommend that assessment of institutional
-              performance should be based on objective evaluation methods,
-              which require the study of patient population outcomes, rather
-              than on subjective methods in which individual cases are
-              reviewed.",
-  journal  = "J. Trauma",
-  volume   =  32,
-  number   =  1,
-  pages    = "45--51",
-  month    =  jan,
-  year     =  1992,
-  language = "en"
-}
-
-#https://journals.lww.com/jtrauma/abstract/1992/01000/identification_of_preventable_trauma_deaths_.10.aspx
-
-@ARTICLE{3,
-  title    = "Patient and process factors associated with opportunities for
-              improvement in trauma care: a registry-based study",
-  author   = "Albaaj, Hussein and Attergrim, Jonatan and Str{\"o}mmer, Lovisa
-              and Brattstr{\"o}m, Olof and Jacobsson, Martin and Wihlke,
-              Gunilla and V{\"a}sterbo, Liselott and Joneborg, Elias and Gerdin
-              W{\"a}rnberg, Martin",
-  abstract = "BACKGROUND: Trauma is one of the leading causes of morbidity and
-              mortality worldwide. Morbidity and mortality review of selected
-              patient cases is used to improve the quality of trauma care by
-              identifying opportunities for improvement (OFI). The aim of this
-              study was to assess how patient and process factors are
-              associated with OFI in trauma care. METHODS: We conducted a
-              registry-based study using all patients between 2017 and 2021
-              from the Karolinska University Hospital who had been reviewed
-              regarding the presence of OFI as defined by a morbidity and
-              mortality conference. We used bi- and multivariable logistic
-              regression to assess the associations between the following
-              patient and process factors and OFI: age, sex, respiratory rate,
-              systolic blood pressure, Glasgow Coma Scale (GCS), Injury
-              Severity Score (ISS), survival at 30 days, highest hospital care
-              level, arrival on working hours, arrival on weekends, intubation
-              status and time to first computed tomography (CT). RESULTS: OFI
-              was identified in 300 (5.8\%) out of 5182 patients. Age, missing
-              Glasgow Coma Scale, time to first CT, highest hospital care level
-              and ISS were statistically significantly associated with OFI.
-              CONCLUSION: Several patient and process factors were found to be
-              associated with OFI, indicating that patients with moderate to
-              severe trauma and those with delays to first CT are at the
-              highest odds of OFI.",
-  journal  = "Scand. J. Trauma Resusc. Emerg. Med.",
-  volume   =  31,
-  number   =  1,
-  pages    = "87",
-  month    =  nov,
-  year     =  2023,
-  keywords = "Acute surgery; Opportunities for improvement; Trauma; Trauma care",
-  language = "en"
-}
-
-#Tidigare publicerad artikel av NOACs
-#Albaaj, H.
-et al.
-Patient and process factors associated with opportunities for improvement in trauma care: a registry based study.
-Scandinavian Journal of Trauma, Resuscitation and Emergency Medicine (2023) 31:87.https://doi.org/10.1186/s13049-023-01157-y.
-
-@ARTICLE{4,
-  title    = "National assessment of opportunities for improvement in
-              preventable trauma deaths: A mixed-methods study",
-  author   = "Kwon, Junsik and Lee, Myeonggyun and Jung, Kyoungwon",
-  abstract = "Trauma is a significant public health issue worldwide,
-              particularly affecting economically active age groups. Quality
-              management of trauma care at the national level is crucial to
-              improve outcomes of major trauma. In Korea, a biennial nationwide
-              survey on preventable trauma death rate is conducted. Based on
-              the survey results, we analyzed opportunities for improving the
-              trauma treatment process. Expert panels reviewed records of 8282
-              and 8482 trauma-related deaths in 2017 and 2019, respectively,
-              identifying 258 and 160 cases in each year as preventable deaths.
-              Opportunities for improvement were categorized into prehospital,
-              interhospital, and hospital stages. Hemorrhage was the primary
-              cause of death, followed by sepsis/multiorgan failure and central
-              nervous system injury. Delayed hemostatic procedures and
-              transfusions were common areas for improvement in hospital stage.
-              Interhospital transfers experienced significant delays in arrival
-              time. This study emphasizes the need to enhance trauma care by
-              refining treatment techniques, centralizing patients in
-              specialized facilities, and implementing comprehensive reviews
-              and performance improvements throughout the patient transfer
-              system. The findings offer valuable insights for addressing
-              trauma care improvement from both clinical and systemic
-              perspectives.",
-  journal  = "Healthcare (Basel)",
-  volume   =  11,
-  number   =  16,
-  month    =  aug,
-  year     =  2023,
-  keywords = "mortality; patient transfer; quality improvement; treatment
-              outcome; wounds and injuries",
-  language = "en"
-}
-
-#https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10454732/pdf/healthcare-11-02291.pdf
-
-```
-
 Methods 
@@
 The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
-The hospital has direct access to radiology, surgery, intensive care and consultants from relevant specialities [6].
+The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6].
 All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
@@
 
-```{r}
-
-@ARTICLE{5,
-  title    = "Trauma center levels defined",
-  author   = "Trauma Center Association of America",
-  URL  = "https://www.traumacenters.org/page/TraumaCenterLevels#Goals",
-  year     =  2014,
-#Hur källhänvisar man webbsida?
-
-@ARTICLE{6,
-  title    = "Traumamanual",
-  author   = "Karolinska Universitetssjukhuset Solna",
-  year     =  2020,
-#Include ISS? And audit filters?
-
-```
-
 Participants
@@
 The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9.
-We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022*.
+We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022 (?)* and who were also admitted to the ICU.
 *We excluded patients who were younger than 15 years and patients who were dead on arrival.*
@@
 Patient and process factors 
-The factors were chosen from the trauma registry, based on the locally used audit filters, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry.
+The factors were chosen from the trauma registry, based on the locally used audit filters ?, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry.
 The categorical factors were *sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated*.
@@
 and 5 p.m., or during a weekend, defined as Saturday or Sunday.
-
-Statistical analysis 
-Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
-The programming language used was R for all analyses [].
-The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
-
-*lägg till?*
-A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above.
-We present sample characteristics using descriptive statistics.
-Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
-
-```{r}
-# LÄNKA till http://www.rstudio.com/ 
-```
 
@@
 -------------------
-*senare*
+Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
+The programming language used was R for all analyses [länk].
+The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
+
+*lägg till?*
+A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above.
+We present sample characteristics using descriptive statistics.
+Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
 

### 2024-01-07 manuscript.Rmd
Comment on manuscript

@@
 ========
+
 Background
@@
 Trauma is a leading cause of mortality and morbidity worldwide.
-A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions.
-Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
+A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions.<!-- This sentence is a bit long, try to break it into two, or use active voice: Multidisciplinary mortality and morbidity review is a ...
+--> Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
 The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
@@
 ============
-Trauma is a leading cause of mortality and morbidity worldwide being the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly [1].
+
+<!-- The structure is generally good.
+I would add some more global data to the beginning, and then try to refer to more studies on potential opportunities for improvement and how they can be addressed.
+Try searching for preventable deaths in trauma patients treated in the ICU as well -->
+
+Trauma is a leading cause of mortality and morbidity worldwide, and the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly<!-- Also in the US?
+I suggest you cite the latest Global Burden of Disease figures, see https://www.thelancet.com/pb-assets/Lancet/gbd/summaries/diseases/injuries.pdf--> [@Frydrych2019]<!-- Use more descriptive keys in bibliography.bib and use them like in this example, i.e.
+within square brackets and preceded by a @-->.
 A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions.
-Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2].
+OFIs provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2].
 Examples of opportunities for improvement include lack of resources and management errors.
@@
 
-According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury.
+According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury<!-- There are many studies on preventable deaths in trauma patients, so that's a good key word to search for -->.
 Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [4].
@@
 We conducted single-centre registry-based retrospective cohort study using data from the trauma registry and trauma care quality database at Karolinska University Hospital in Solna.
-The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
+<!-- This can be removed as it should be covered in other sections: The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2022.
 The outcome of the selected patients was OFI identified by the multidisciplinary review board and then recorded in the trauma care quality database.
@@
 A 5% significance level and 95% confidence levels will be used.
+-->
 
@@
 -------
-The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
+The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], and is designated to receive all patients in Stockholm with major or suspected major trauma.
 The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6].
@@
 ---------------------------------------
-Study outcome
+
+### Study outcome
+
 The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors.
-*vilka OFIs ska läggas till för att analyseras?*
+*vilka OFIs ska läggas till för att analyseras?*<!-- You can keep it like this for now -->
 
-Patient and process factors 
+### Patient and process factors
+
 The factors were chosen from the trauma registry, based on the locally used audit filters ?, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry.
@@
 ----------
+
+<!-- Read up on the sample size requirements of logistic regression.
+You will come across for example the 10 events per variable rule, which is sometimes used as a rule of thumb to decide the size of the model in comparison to available data.
+You can also read up on the concept of power.
+-->
+
 *senare*
@@
 
+<!-- Age, ISS, time from arrival to CT, systolic blood pressure, respiratory rate and GCS score can all be considered quantitative variables. Describe how they were handled in the analyses. -->
+
 <!-- Explain how quantitative variables were handled in the
@@
 -------------------
-Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
-The programming language used was R for all analyses [länk].
-The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
-
-*lägg till?*
+We used R for all statistical analyses [länk].
 A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above.
 We present sample characteristics using descriptive statistics.
-Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
+The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database <!-- This can be moved to the bias section -->.
+<!-- This is where you write about the logistic regression --> Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
 

### 2024-01-11 manuscript.Rmd
commit 11/1

--- 2024-01-05 manuscript.Rmd
@@
 Trauma is a leading cause of mortality and morbidity worldwide.
-A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions.
+A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review.
+The purpose of these is to identify opportunities for improvement (OFI), which are necessary in order to implement corrective actions.
 Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
@@
 -------
-This was a registry-based study using all patients between 2014 and 2022 from the Karoliniska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.
+This was a registry-based study using all patients between 2014 and 2022 from the Karolinska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.
 Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...
@@
 ============
-Trauma is a leading cause of mortality and morbidity worldwide being the leading cause of death among people under 44 years old in the United States and the fourth leading cause of death in the elderly [1].
-A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review to identify opportunities for improvement (OFI) in order to implement corrective actions.
-Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner rather than only subjective methods where individual cases are reviewed [2].
-Examples of opportunities for improvement include lack of resources and management errors.
+Trauma is a leading cause of mortality and morbidity worldwide.
+According to WHO, injury-related deaths constitute 4% of all deaths and take the lives of 4,4 million people yearly worldwide - more than HIV/AIDS, malaria, maternal mortality and tuberculosis combined.
+The top 3 of 5 causes of death in people aged 5-29 are injury-related [1,2].
+Trauma accounts for one-tenth of disability-adjusted life-years (DALYs), most of the burden in low and middle income countries (LMIC) [3, 4].
+Trauma costs approximately 3% of GDP yearly globally [5].
 
+A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review.
+The purpose of these is to identify opportunities for improvement (OFI).
+Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner [6].
 Albaaj et al.
 recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna.
-The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [3].
+The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [7].
 Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
 
-According to a study on the Korean trauma registry the primary cause of death after trauma was haemorrhage, followed by sepsis/multiorgan failure and central nervous system injury.
-Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [4].
-These are examples of factors that could be important in care on a higher hospital level such as the ICU.
+There are many studies on preventable deaths in trauma patients treated in the ICU both regarding to the medical aspect as well as the human aspect.
+The primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury.
+Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [8].
+The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill.
+The OFIs frequently involved the decision between radiology, surgery and further investigation [9].
+
+The University of Pennsylvania ICUs uses telemedicine and telemonitoring to review critical patient events.
+The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique.
+They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [10].
+Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items.
+This points at the importance of communication and handling of critical information at handoffs of care [11].
+From the studies mentioned above some OFIs related to the ICU can be identified: decision-making, access to resources and communication; however, there is still much left to discover.
 The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
@@
 ------------
-We conducted single-centre registry-based retrospective cohort study using data from the trauma registry and trauma care quality database at Karolinska University Hospital in Solna.
-The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.
-The outcome of the selected patients was OFI identified by the multidisciplinary review board and then recorded in the trauma care quality database.
-From 2017 the trauma care quality database includes all patients in the trauma registry as well as the results of the mortality and morbidity review.
-This project will use these two databases and assess what types of OFI occurred in ICU patients and how different patient factors; such as age, sex, mechanism of injury and injury severity, are associated with OFI using logistic regression.
-A 5% significance level and 95% confidence levels will be used.
+?
 
@@
 -------
-The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
-The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6].
+The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [12], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
+The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [13].
 All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
@@
 The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9.
-We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and *June 1, 2022 (?)* and who were also admitted to the ICU.
-*We excluded patients who were younger than 15 years and patients who were dead on arrival.*
+We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and June 1, 2022 and who were also admitted to the ICU.
+We excluded patients who were younger than 15 years and patients who were dead on arrival.
 
@@
 Study outcome
-The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors.
-*vilka OFIs ska läggas till för att analyseras?*
+-------------
+The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors...
 
 Patient and process factors 
-The factors were chosen from the trauma registry, based on the locally used audit filters ?, *standard epidemiological factors* and factors registered in the Swedish National Trauma Registry.
-The categorical factors were *sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated*.
+---------------------------
+The factors were chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry.
+The categorical factors were sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated.
 The continuous factors age and Injury Severity Score (ISS) were also included.
@@
 ----
-*senare*
-
-<!-- Describe any efforts to address potential sources of bias -->
+We used R for all statistical analyses [].
+A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score...
+We present sample characteristics using descriptive statistics.
+The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
 
@@
 ----------
-*senare*
-
 <!-- Explain how the study size was arrived at -->
@@
 ----------------------
-*senare*
-
 <!-- Explain how quantitative variables were handled in the
@@
 -------------------
+Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
 Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
-The programming language used was R for all analyses [länk].
-The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
-
-*lägg till?*
-A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score as described above.
-We present sample characteristics using descriptive statistics.
-Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
 

### 2024-01-11 main.R
11/1

@@ -1,26 +1,0 @@
-## Welcome!
-
-## This is your project's main script file and together with
-## manuscript.Rmd it provides and entry point for you and other people
-## coming to the project. The code in this file should give an outline
-## of the different steps conducted in your study, from importing data
-## to producing results.
-
-## This file should be relatively short, and most of the heavy
-## lifting should be done by specialised functions. These functions
-## live in the folder functions/ and you create a new function using
-## create_function().
-
-## Feel free to remove this introductory text as you get started.
-
-## Source all functions (if you tick the box "Source on save" in
-## RStudio functions will be automatically sourced when you save
-## them). They all need to be sourced however when you compile your
-## manuscript file or run this file as a job, as that happens in a
-## clean R session.
-noacsr::source_all_functions()
-
-## Import data
-data <- import_data(test = TRUE)
-
-## Whatever you do next, maybe clean data?

### 2024-01-11 manuscript.Rmd
11/1

@@
 
+Setting
+-------
+The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], and is designated to receive all patients in Stockholm with major or suspected major trauma.
+The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6].
+All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
+The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks.
+Examples of OFI include the need for better organization or more senior members taking part in the trauma care.
+The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.
+
 Participants
@@
 ---------------------------------------
-Study outcome
--------------
+
+### Study outcome
 The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors...
 
-Patient and process factors 
----------------------------
+### Patient and process factors 
 The factors were chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry.
@@
 ----------
+
+<!-- Read up on the sample size requirements of logistic regression.
+You will come across for example the 10 events per variable rule, which is sometimes used as a rule of thumb to decide the size of the model in comparison to available data.
+You can also read up on the concept of power.
+-->
+
 <!-- Explain how the study size was arrived at -->
@@
 ----------------------
+*senare*
+
+<!-- Age, ISS, time from arrival to CT, systolic blood pressure, respiratory rate and GCS score can all be considered quantitative variables. Describe how they were handled in the analyses. -->
+
 <!-- Explain how quantitative variables were handled in the

### 2024-01-11 main.R

@@ -1,0 +1,46 @@
+## Welcome!
+
+## This is your project's main script file and together with
+## manuscript.Rmd it provides and entry point for you and other people
+## coming to the project. The code in this file should give an outline
+## of the different steps conducted in your study, from importing data
+## to producing results.
+
+## Feel free to remove this introductory text as you get started.
+
+## Before you start writing code you need to install some packages.
+## Run the following code to install the packages you need:
+install.packages(devtools)
+devtools::install_github("martingerdin/noacsr")
+devtools::install_github("martingerdin/rofi")
+
+## You can remove the three lines above once the packages have been
+## installed, or put ## in front of each line to comment them out, if
+## you want to keep them for future reference.
+
+## Load packages
+library(dotenv)
+library(noacsr)
+library(rofi)
+noacsr::source_all_functions()
+
+## Import data
+data <- import_data(test = TRUE)
+
+## Merge data
+merged.data <- merge_data(data)
+
+## Add outcome variable ofi
+merged.data$ofi <- create_ofi(merged.data)
+
+## This file should be relatively short, and most of the heavy
+## lifting should be done by specialised functions. These functions
+## live in the folder functions/ and you create a new function using
+## create_function().
+
+## Source all functions (if you tick the box "Source on save" in
+## RStudio functions will be automatically sourced when you save
+## them). They all need to be sourced however when you compile your
+## manuscript file or run this file as a job, as that happens in a
+## clean R session.
+

### 2024-01-11 manuscript.Rmd

@@
 
-Setting
--------
-The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [5], and is designated to receive all patients in Stockholm with major or suspected major trauma.
-The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [6].
-All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
-The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks.
-Examples of OFI include the need for better organization or more senior members taking part in the trauma care.
-The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.
-
 Participants
@@
 ---------------------------------------
-
 ### Study outcome

### 2024-01-20 manuscript.Rmd
commit 20/1

--- 2024-01-11 manuscript.Rmd
@@
 The purpose of these is to identify opportunities for improvement (OFI), which are necessary in order to implement corrective actions.
-Many patients with severe trauma is admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
+Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
 The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
@@
 This was a registry-based study using all patients between 2014 and 2022 from the Karolinska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.
-Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: ...
+Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: gender, age, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT), if the patient was intubated and Injury Severity Score (ISS).
 
@@
 
+Abbreviations
+=============
+CT = computed tomography 
+GCS = Glasgow Coma Scale
+GOS = Glasgow Outcome Score
+ISS = Injury Severity Score 
+NISS = New Injury Severity Score 
+OFI = Opportunity For Improvement 
+RTS = Revised Trauma Score 
+TRISS = Trauma and Injury Severity Score
+
 Introduction
 ============
-Trauma is a leading cause of mortality and morbidity worldwide.
+Trauma — the clinical entity of injury and the body's associated response — is a leading cause of mortality and morbidity worldwide.
 According to WHO, injury-related deaths constitute 4% of all deaths and take the lives of 4,4 million people yearly worldwide - more than HIV/AIDS, malaria, maternal mortality and tuberculosis combined.
-The top 3 of 5 causes of death in people aged 5-29 are injury-related [1,2].
-Trauma accounts for one-tenth of disability-adjusted life-years (DALYs), most of the burden in low and middle income countries (LMIC) [3, 4].
-Trauma costs approximately 3% of GDP yearly globally [5].
+The top 3 of 5 causes of death in people aged 5-29 are injury-related; road traffic injuries, homicide and suicide.
+Each year, tens of millions of people suffer non-fatal injuries leading to treatment at hospitals in emergency departments and acute care visits.
+These injuries often lead to temporary or permanent disability and the need for medical and mental rehabilitation.
+Trauma exposure at a young age increases the risk for mental illness and suicide leading to increased risk of an unhealthy lifestyle associated with smoking, alcohol and substance abuse, chronic disease and cancer as well as societal problems such as crime, poverty and violence.
+The burden of disease from injuries vary depending on certain conditions; socioeconomic factors, gender, age and country.
+Men suffer from injuries more often than women.
+Across all age groups worldwide, the top three leading causes of injury among males are road traffic injuries, homicide and suicide; whereas the top three leading causes of injury among women are road traffic injuries, falls and suicide [1,2].
 
-A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review.
-The purpose of these is to identify opportunities for improvement (OFI).
-Identifying OFIs will provide guidance as to what to focus specific efforts on and is a way to assess trauma patient cases in a more objective manner [6].
+According to the annual report from the Swedish Trauma registry (SweTrau) 2022, trauma was more common among men than women in Sweden — 65,3% among men compared to 34,7% among women.
+The most affected age group was the working population.
+The leading causes of trauma included; road traffic injuries, falls, weapons and stabs and blows.
+Previous years, road traffic injuries have been the leading cause of injury.
+However, since 2022 road traffic injuries and falls account for about the same proportion of mechanisms of injury.
+There are several variables that can be studied to assess primary patient care, for example time to first CT and acute interventions.
+Median time to first CT in the university hospitals of Sweden was 30 minutes in unconscious patients with GCS < 9.
+Acute interventions is defined as interventions performed immediately or within 24 hours after arrival to a hospital.
+The most common acute interventions in Swedish hospitals were: thoracic drains, wound revisions and fracture surgery followed by laparotomy with haemostasis, radiological interventions such as embolisation and stents, and thoracotomy [3].
+
 Albaaj et al.
-recently analyzed OFI in relation to different patient and process factors based on data from the trauma registry of Karolinska Institutet Solna.
-The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [7].
+recently analyzed OFI in relation to different patient and process factors based on data from the Swedish trauma registry and data from the Karolinska University Hospital  in Solna.
+The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [4].
+According to SweTrau 2022 41,1% of patients with New Injury Severity Score (NISS) > 15 who were at a university hospital were admitted to the ICU.
+The proportion of severely injured admitted to the ICU has decreased in recent years, however; patients with NISS > 15 still constitute a large proportion of the trauma patients treated in the ICU.
+A way to measure functional level after hospital discharge is Glasgow Outcome Scale (GOS).
+According to GOS, among patients with NISS > 15 17% were deceased, 20% suffered severe disability, 49% suffered moderate disability and 14% had good recovery in the university hospitals of Sweden [3].
+
+For all injuries, quality hospital care can reduce the amount of short- and long-term disability.
+Therefore, improving planning, access and organisation of trauma care systems including pre-hospital care, interhospital transportation, telecommunication and in-hospital care all play a central role in reducing overall impact after trauma [1].
+Disability-adjusted life-years (DALYs) is a way to measure post-discharge quality of life.
+Trauma accounts for one-tenth of DALYs, most of the burden in low and middle income countries (LMIC) [5, 6].
+Trauma costs approximately 3% of gross domestic product (GDP) yearly globally, as it mostly affects the working population [7].
+
+Trauma quality improvement programs
+-----------------------------------
+Many initiatives for quality improvement programs have been initiated in trauma and critical care based on the paradigm of Donabedian [8], stating that implementing proven structures and processes of care is the most effective way to improve outcomes.
+One example of a structure and process intervention in the intensive care unit is highly-staffed ICUs that has been associated with improved outcomes, as well as bundles to recognize and treat common conditions such as those for treating sepsis and ventilator-associated pneumonia [9].
+The intention of quality improvement is objective improvement in regards to preventability.
+A cornerstone in developing the trauma quality improvement programs is therefore multidisciplinary mortality and morbidity review.
+The purpose of these is to discuss preventable deaths and identifying opportunities for improvement (OFI) in trauma care regarding structure as well as clinical processes.
+This way, identifying OFIs will provide guidance as to what to focus specific efforts on.
+There are several other methods to improve trauma care.
+However, mortality and morbidity review has been shown to be a key method of addressing all necessary components of trauma improvement [10].
+
+Trauma audit filters are descriptions of specific actions that should be taken, time frames in which tests or treatments should be provided, or outcomes that are expected to occur in injured patients.
+Assessing patients whose care falls outside the frames of these audit filters and providing feedback to the clinicians involved in the patient's care provides means of correcting errors and improving future trauma care.
+The audit filters were introduced by the American College of Surgeons 1995; however, they have been criticized as they may not be directly linked to meaningful clinical or patient-centered outcomes [11].
+At the Karolinska University Hospital in Solna local audit filters are used, shown in Table 1.
+These local audit filters include: systolic blood pressure under 90, GCS < 9 and not intubated, ISS > 15 but not admitted to the ICU, ISS > 15 and no trauma team activation, no anticoagulation treatment within 72 h after traumatic brain, injury time to acute intervention more than 60 minutes, time to computed tomography (CT) more than 30 minutes, liver or spleen injury, cardiopulmonary resuscitation with thoracotomy, massive transfusion and death within 30 days after trauma.
+
+Injury score systems
+--------------------
+When a patient first arrives at the hospital, injury system scores are used to assess how urgent the need of interventions to a patient is.
+There are several injury score systems: Injury Severity Score (ISS), New Injury Severity Score (NISS), Revised Trauma Score (RTS) and Trauma and Injury Severity Score (TRISS).
+
+ISS is an anatomical scoring system providing an overall score for patients with multiple injuries.
+Each injury is assigned an abbreviated injury scale (AIS) score and is allocated to one of six body regions.
+The three most severely injured body regions have their scores squared and added together to produce an ISS score between 0 and 75.
+If an injury is assigned an AIS of 6 the ISS score is automatically assigned to 75.
+Major trauma is considered when ISS > 15.
+The major drawback of ISS is that it only considers one injury for each body region.
+Therefore, if several severe injuries are in the same body region, they are not accounted for.
+NISS counters the problem of ISS being the sum of the square of the three most severe injuries, regardless of body region injured.
+NISS is therefore equal to or higher than ISS.
+
+RTS is a physiological scoring system scored from the data obtained in the first set of datas obtained on the patient and consists of Glasgow Coma Scale (GCS), respiratory rate and systolic blood pressure.
+RTS is heavily weighted towards GCS compared to respiratory rate and systolic blood pressure.
+It has been demonstrated to show accuracy at predicting death.
+
+TRISS has a maximum prediction in outcome when compared with the other scores.
+The score determines the probability of survival based on a formula consisting of ISS, RTS and the patient's age.
+TRISS is therefore both an anatomical and physiological scoring system [12].
+
+Preventable deaths in patients admitted to the ICU
+--------------------------------------------------
+There are many studies on preventable deaths in trauma patients treated in the ICU both regarding the medical aspect as well as the human aspect.
+According to a study from Japan on the medical aspect, the primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury.
+Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [13].
+The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill.
+The OFIs frequently involved the decision between radiology, surgery and further investigation [14].
+
+A study from The University of Pennsylvania looked at the human aspect of OFIs in the ICU.
+The ICU of this hospital uses telemedicine and telemonitoring to review critical patient events.
+The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique.
+They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [15].
+Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items.
+This points at the importance of communication and handling of critical information at handoffs of care [16].
+
+Preventability in studies is often defined as an event that would not have occurred had the patient received ordinary standards of care.
+Evaluating preventable deaths in patients admitted to the ICU can be a challenge as the patients have extensive medical history and it is harder to understand what the standard of care is for those patients.
+Furthermore, even though these patients receive the best of care, these patients still have a higher risk of developing complications or dying.
+Some might even be in such a bad condition that they die just when they have been admitted to the ICU from the ward or emergency care [9].
+
+Aim
+---
 Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
-
-There are many studies on preventable deaths in trauma patients treated in the ICU both regarding to the medical aspect as well as the human aspect.
-The primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury.
-Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [8].
-The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill.
-The OFIs frequently involved the decision between radiology, surgery and further investigation [9].
-
-The University of Pennsylvania ICUs uses telemedicine and telemonitoring to review critical patient events.
-The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique.
-They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [10].
-Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items.
-This points at the importance of communication and handling of critical information at handoffs of care [11].
-From the studies mentioned above some OFIs related to the ICU can be identified: decision-making, access to resources and communication; however, there is still much left to discover.
-The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
+This being due to the patients admitted to the ICU being more complex.
+Identifying OFIs in this group can therefore be more of a challenge.
+There is existing studies on different aspects of trauma care improvement, but studies focusing mainly on OFI in the ICU are few.
+The aim of this study is therefore to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
 
@@
 ------------
-?
+We conducted a retrospective single-centre cohort study based on data from the Karolinska University Hospital in Solna and from the Swedish Trauma registry.
 
@@
 -------
-The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [12], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
-The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [13].
+The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [17], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
+The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [18].
 All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
@@
 ---------------------------------------
-Study outcome
--------------
-The study outcome will be on the presence of OFI in adult patients treated in the ICU focusing on different types of opportunities for improvement including delayed treatment, judgement error and technical errors...
+### Study outcome
+The study outcome will be on the presence of OFI in adult patients treated in the ICU based on patient and process factors.
 
-Patient and process factors 
----------------------------
-The factors were chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry.
+### Patient and process factors 
+The quantitative variables were factors chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry.
 The categorical factors were sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated.
@@
 
-Bias 
-----
-We used R for all statistical analyses [].
-A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score...
+Bias/ethical considerations
+---------------------------
+We used R for all statistical analyses [19].
+A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score.
 We present sample characteristics using descriptive statistics.
-The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database.
+The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database in order to protect personal numbers and their associated information.
 
@@
 ----------
-<!-- Explain how the study size was arrived at -->
-
-Quantitative variables
-----------------------
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
+In the data there are 167 outcomes. Will later explain how the study size was arrived at.
 
@@
 <!-- 
-
 (a) Describe all statistical methods, including those used to control
 for confounding
-
 (b) Describe any methods used to examine subgroups and interactions
-
 (c) Explain how missing data were addressed 
-
 (d) Cohort study: If applicable, explain how loss to follow-up was addressed 
-
 (e) Describe any sensitivity analyses
@@
 =======
+Table 1: deskriptiv tabell, stratifierad från utfallet, förslag är att använda: https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html
 
-<!--
+Table 2: huvudresultat, förslag att använda: https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html 
+presentera ojusterad (univariable logistisk regression) och justerad (multivariable log reg)
+
+Tillse korrekt presentation av resultaten i textformat: https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html?q=tbl%20_%20re#inline_text_tbl_regression
+
+Ytterligare tabeller, audit filtren? En flow chart för inklusion/exklusion?
 
@@
 
-(a) Report numbers of individuals at each stage of study—eg numbers
+<!-- (a) Report numbers of individuals at each stage of study—eg numbers
 potentially eligible, examined for eligibility, confirmed eligible,
@@
 
+-->
+
 Descriptive data
@@
 
-(a) Give characteristics of study participants (eg demographic,
+<!-- (a) Give characteristics of study participants (eg demographic,
 clinical, social) and information on exposures and potential
@@
 amount)
+-->
 
@@
 
-Cohort study — Report numbers of outcome events or summary measures
+<!-- Cohort study — Report numbers of outcome events or summary measures
 over time
+-->
 
@@
 
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
+<!-- (a) Give unadjusted estimates and, if applicable, confounder-adjusted
 estimates and their precision (eg, 95% confidence interval). Make
@@
 absolute risk for a meaningful time period
+-->
 
@@
 
-Report other analyses done—eg analyses of subgroups and interactions,
+<!--Report other analyses done—eg analyses of subgroups and interactions,
 and sensitivity analyses 

### 2024-01-20 main.R

@@ -1,37 +1,9 @@
-## Welcome!
-
+## INTRUCTIONS 
 ## This is your project's main script file and together with
 ## manuscript.Rmd it provides and entry point for you and other people
 ## coming to the project. The code in this file should give an outline
 ## of the different steps conducted in your study, from importing data
 ## to producing results.
-
-## Feel free to remove this introductory text as you get started.
-
-## Before you start writing code you need to install some packages.
-## Run the following code to install the packages you need:
-install.packages(devtools)
-devtools::install_github("martingerdin/noacsr")
-devtools::install_github("martingerdin/rofi")
-
-## You can remove the three lines above once the packages have been
-## installed, or put ## in front of each line to comment them out, if
-## you want to keep them for future reference.
-
-## Load packages
-library(dotenv)
-library(noacsr)
-library(rofi)
-noacsr::source_all_functions()
-
-## Import data
-data <- import_data(test = TRUE)
-
-## Merge data
-merged.data <- merge_data(data)
-
-## Add outcome variable ofi
-merged.data$ofi <- create_ofi(merged.data)
 
 ## This file should be relatively short, and most of the heavy
 ## lifting should be done by specialised functions. These functions
@@ -43,3 +15,138 @@ merged.data$ofi <- create_ofi(merged.data)
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
+
+
+
+## install.packages(devtools)
+## devtools::install_github("martingerdin/noacsr")
+## devtools::install_github("martingerdin/rofi")
+
+## Load packages
+library(dotenv)
+library(noacsr)
+library(rofi)
+noacsr::source_all_functions()
+
+## Import data
+data <- import_data(test = TRUE)
+
+## Merge data
+merged.data <- merge_data(data, test = TRUE)
+
+## Add outcome variable ofi
+## merged.data$ofi <- create_ofi(merged.data)
+## The function "create_ofi" creates OFI variable which is either yes or no. 
[…]

### 2024-01-31 manuscript.Rmd
31/1

@@
 csl: vancouver.csl
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+# Abstract
+
+## Background
+
+Trauma is a leading cause of mortality and morbidity worldwide. A
+cornerstone in developing the trauma quality improvement programs is
+multidisciplinary mortality and morbidity review. The purpose of these
+is to identify opportunities for improvement (OFI), which are necessary
+in order to implement corrective actions. Many patients with severe
+trauma are admitted to the intensive care unit (ICU), but little is
+known about opportunities for improvement in this group of patients. The
+aim of this study is to characterise OFI in adult trauma patients
+admitted to the ICU and assess how patient factors are associated with
+OFI in these patients.
+
+## Methods
+
+This was a registry-based study using all patients between 2014 and 2022
+from the Karolinska University Hospital in Solna who had been admitted
+to the ICU and who were reviewed regarding the presence of OFI by the
+mortality and morbidity conference. Bi-variable and multivariable
+logistic regression was used to assess the following patient and process
+factors and OFI among this group of patients: gender, age, survival
+after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic
+blood pressure, working hours, weekend, time from arrival at the
+hospital until first computed tomography (CT), if the patient was
+intubated and Injury Severity Score (ISS).
+
+## Results
+
+## Conclusion
+
+# Abbreviations
+
+CT = computed tomography
+
+GCS = Glasgow Coma Scale
+
+GOS = Glasgow Outcome Score
+
+ISS = Injury Severity Score
+
+NISS = New Injury Severity Score
+
+OFI = Opportunity For Improvement
+
+RTS = Revised Trauma Score
+
+TRISS = Trauma and Injury Severity Score
+
+# Introduction
+
+Trauma --- the clinical entity of injury and the body's associated
+response --- is a leading cause of mortality and morbidity worldwide.
+According to WHO, injury-related deaths constitute 4% of all deaths and
+take the lives of 4,4 million people yearly worldwide - more than
+HIV/AIDS, malaria, maternal mortality and tuberculosis combined. The top
+3 of 5 causes of death in people aged 5-29 are injury-related; road
+traffic injuries, homicide and suicide. Each year, tens of millions of
+people suffer non-fatal injuries leading to treatment at hospitals in
+emergency departments and acute care visits. These injuries often lead
+to temporary or permanent disability and the need for medical and mental
+rehabilitation. Trauma exposure at a young age increases the risk for
+mental illness and suicide leading to increased risk of an unhealthy
+lifestyle associated with smoking, alcohol and substance abuse, chronic
+disease and cancer as well as societal problems such as crime, poverty
+and violence. The burden of disease from injuries vary depending on
+certain conditions; socioeconomic factors, gender, age and country. Men
+suffer from injuries more often than women. Across all age groups
+worldwide, the top three leading causes of injury among males are road
+traffic injuries, homicide and suicide; whereas the top three leading
+causes of injury among women are road traffic injuries, falls and
+suicide [@1,@2].
+
+According to the annual report from the Swedish Trauma registry
+(SweTrau) 2022, trauma was more common among men than women in Sweden
+--- 65,3% men compared to 34,7% women. The most affected age group was
+the working population. The leading causes of trauma included; road
+traffic injuries, falls, weapons and stabs and blows. Previous years,
+road traffic injuries have been the leading cause of injury. However,
+since 2022 road traffic injuries and falls account for about the same
+proportion of mechanisms of injury. There are several variables that can
+be studied to assess primary patient care, for example time to first CT
+and acute interventions. Median time to first CT at the university
+hospitals of Sweden was 30 minutes in unconscious patients with GCS \<
+9. Acute interventions is defined as interventions performed immediately
+or within 24 hours after arrival to a hospital. The most common acute
+interventions in Swedish hospitals were: thoracic drains, wound
+revisions and fracture surgery followed by laparotomy with haemostasis,
+radiological interventions such as embolisation and stents, and
+thoracotomy [@3].
+
+Albaaj et al. recently analyzed opportunities for improvement (OFI) in
+relation to different patient and process factors based on data from the
+Swedish trauma registry and data from the Karolinska University Hospital
+in Solna. The study indicated that patients with moderate to severe
+trauma are at the highest odds of OFI [@4]. According to SweTrau 2022
+41,1% of patients with New Injury Severity Score (NISS) \> 15 who were
+at a university hospital were admitted to the ICU. The proportion of
+severely injured admitted to the ICU has decreased in recent years,
+however; patients with NISS \> 15 still constitute a large proportion of
+the trauma patients treated in the ICU. A way to measure functional
+level after hospital discharge is Glasgow Outcome Scale (GOS). According
+to GOS, among patients with NISS \> 15 17% were deceased, 20% suffered
+severe disability, 49% suffered moderate disability and 14% had good
+recovery in the university hospitals of Sweden [@3].
+
+For all injuries, quality hospital care can reduce the amount of short-
+and long-term disability. Therefore, improving planning, access and
+organisation of trauma care systems including pre-hospital care,
+interhospital transportation, telecommunication and in-hospital care all
+play a central role in reducing overall impact after trauma [@1].
+Disability-adjusted life-years (DALYs) is a way to measure
+post-discharge quality of life. Trauma accounts for one-tenth of DALYs,
+most of the burden in low and middle income countries (LMIC) [@5, @6].
+Trauma costs approximately 3% of gross domestic product (GDP) yearly
+globally, as it mostly affects the working population [@7].
+
+## Trauma quality improvement programs
+
+Many initiatives for quality improvement programs have been seen in
+trauma and critical care based on the paradigm of Donabedian [@8],
+stating that implementing proven structures and processes of care is the
+most effective way to improve outcomes. One example of a structure and
+process intervention in the intensive care unit is highly-staffed ICUs
+that has been associated with improved outcomes, as well as bundles to
+recognize and treat common conditions such as those for treating sepsis
+and ventilator-associated pneumonia [@9]. The intention of quality
+improvement is objective improvement in regards to preventability. A
+cornerstone in developing the trauma quality improvement programs is
+therefore multidisciplinary mortality and morbidity review. The purpose
+of these is to discuss preventable deaths and identifying OFIs in trauma
+care regarding structure as well as clinical processes. This way,
+identifying OFIs will provide guidance as to what to focus specific
+efforts on. There are several other methods to improve trauma care.
+However, mortality and morbidity review has been shown to be a key
+method of addressing all necessary components of trauma improvement
+[@10].
+
+Trauma audit filters are descriptions of specific actions that should be
+taken, time frames in which tests or treatments should be provided, or
+outcomes that are expected to occur in injured patients. Assessing
+patients whose care falls outside the frames of these audit filters and
+providing feedback to the clinicians involved in the patient's care
+provides means of correcting errors and improving future trauma care.
+The audit filters were introduced by the American College of Surgeons
+1995; however, they have been criticized as they may not be directly
+linked to meaningful clinical or patient-centered outcomes [@11]. At the
+Karolinska University Hospital in Solna local audit filters are used,
+shown in Table 1. These local audit filters include: systolic blood
+pressure under 90, GCS \< 9 and not intubated, ISS \> 15 but not
+admitted to the ICU, ISS \> 15 and no trauma team activation, no
+anticoagulation treatment within 72 h after traumatic brain injury, time
+to acute intervention more than 60 minutes, time to CT more than 30
+minutes, liver or spleen injury, cardiopulmonary resuscitation with
+thoracotomy, massive transfusion and death within 30 days after trauma.
+
+## Injury score systems
+
+When a patient first arrives at the hospital, injury score systems are
+used to assess how urgent the need of interventions to a patient is.
+There are several injury score systems: Injury Severity Score (ISS), New
+Injury Severity Score (NISS), Revised Trauma Score (RTS) and Trauma and
+Injury Severity Score (TRISS).
+
+ISS is an anatomical scoring system providing an overall score for
+patients with multiple injuries. Each injury is assigned an abbreviated
+injury scale (AIS) score and is allocated to one of six body regions.
+The three most severely injured body regions have their scores squared
+and added together to produce an ISS score between 0 and 75. If an
+injury is assigned an AIS of 6 the ISS score is automatically assigned
+to 75. Major trauma is considered when ISS \> 15. The major drawback of
+ISS is that it only considers one injury for each body region.
+Therefore, if several severe injuries are in the same body region, they
+are not accounted for. NISS counters the problem of ISS being the sum of
+the squares of the three most severe injuries, regardless of body region
+injured. NISS is therefore equal to or higher than ISS.
+
+RTS is a physiological scoring system scored from the data obtained in
+the first set of data obtained on the patient and consists of Glasgow
+Coma Scale (GCS), respiratory rate and systolic blood pressure. RTS is
+heavily weighted towards GCS compared to respiratory rate and systolic
+blood pressure. It has been demonstrated to show accuracy at predicting
+death.
+
+TRISS has a maximum prediction in outcome when compared with the other
+scores. The score determines the probability of survival based on a
+formula consisting of ISS, RTS and the patient's age. TRISS is therefore
+both an anatomical and physiological scoring system [@12].
+
+## Preventable deaths in patients admitted to the ICU
+
+There are many studies on preventable deaths in trauma patients treated
+in the ICU both regarding the medical aspect as well as the human
+aspect. According to a study from Japan on the medical aspect, the
+primary medical causes of death after trauma are haemorrhage,
+sepsis/multiorgan failure and central nervous system injury. Delayed
+hemostatic procedures and transfusions were common areas for improvement
+in the hospital stage [@13]. The most prominent OFI in regards to
+haemorrhage was decision-making compared to errors in technical skill.
+The OFIs frequently involved the decision between radiology, surgery and
+further investigation [@14].
+
+A study from The University of Pennsylvania looked at the human aspect
+of OFIs in the ICU. The ICU of this hospital uses telemedicine and
+telemonitoring to review critical patient events. The areas of
+improvement they identified were: team dynamics and communication, use
+of best practices, increased and standardized access to emergency
+resources and improvement of procedural technique. They recommend other
+ICUs with telemedicine capacity to adopt a similar strategy in order to
+present OFIs and educational discussions [@15]. Another study looked at
+how critical information was handled in the ICU and found that
+laboratory values and test results were the most frequently lost items.
+This points at the importance of communication and handling of critical
+information at handoffs of care [@16].
+
+Preventability in studies is often defined as an event that would not
+have occurred had the patient received ordinary standards of care.
+Evaluating preventable deaths in patients admitted to the ICU can be a
+challenge as the patients have extensive medical history and it is
+harder to understand what the standard of care is for those patients.
+Furthermore, even though these patients receive the best of care, these
+patients still have a higher risk of developing complications or dying.
+Some might even be in such a bad condition that they die just when they
+have been admitted to the ICU from the ward or emergency care [@9].
+
+## Aim
+
+Many patients with severe trauma are admitted to the intensive care unit
+(ICU), but little is known about opportunities for improvement in this
+group of patients. This being due to the patients admitted to the ICU
+being more complex. Identifying OFIs in this group can therefore be more
+of a challenge. There are existing studies on different aspects of
+trauma care improvement, but studies focusing mainly on OFI in the ICU
+are few. The aim of this study is therefore to characterise OFI in adult
+trauma patients admitted to the ICU and assess how patient factors are
+associated with OFI in these patients.
+
+# Methods
+
+## Study design
+
+We conducted a retrospective single-centre cohort study based on data
+from the Karolinska University Hospital in Solna and from the Swedish
+Trauma registry. We used R for all statistical analyses [@19]. A
+complete case analysis was conducted after handling missing values in
+systolic blood pressure, respiratory rate and GCS score. We present
+sample characteristics using descriptive statistics.
+
+## Setting
+
+The Karolinska University Hospital in Solna is a level 1 trauma centre
+based on the definition by the American College of Surgeons (ACS)
+Committee on Trauma [@17], which means that all patients in Stockholm
+with severe or suspected severe trauma are sent there. The hospital has
+direct access to radiology, surgery, emergency medicine, intensive care
+and consultants from relevant specialities [@18]. All patients are
+included in a mortality and morbidity screening combining individual
+review and specific audit filters by specialized nurses. The patients
+who are identified as having higher probability for OFI are discussed in
+a multidisciplinary conference held every 6-8 weeks. Examples of OFI
+include the need for better organization or more senior members taking
+part in the trauma care. The presence or absence of OFI is decided among
+the participants of the conference and then recorded in the trauma care
+quality database.
+
+## Participants
+
+The trauma registry includes all patients admitted with trauma team
+activation, regardless of Injury Severity Score (ISS), as well as
+patients admitted without trauma team activation but found to have an
+ISS of more than 9. We included all patients who had been included in
+the morbidity and mortality screening process between January 1, 2014
+and June 1, 2022 and who were also admitted to the ICU. We excluded
+patients who were younger than 15 years and patients who were dead on
+arrival.
+
+## Variables and data sources/measurements
+
+### Study outcome
+
+The study outcome will be on the presence of OFI in adult patients
+treated in the ICU based on patient and process factors.
+
+### Patient and process factors
+
+The quantitative variables were factors chosen from the trauma registry,
+based on the locally used audit filters, standard epidemiological
+factors and factors registered in the Swedish Trauma Registry. The
+categorical factors were sex, survival after 30 days, GCS, respiratory
+rate, systolic blood pressure, working hours, weekend, time from arrival
+at the hospital until first CT, if the patient was intubated, age and
+Injury Severity Score (ISS). Working hours was defined as between 8.00
+a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.
+
+## Study size
+
+There were ... patients in the trauma registry. ... with trauma team
+activation regardless of ISS score, ... patients admitted without trauma
+team activation but with ISS score \> 9. .... patients were included in
+the morbidity and mortality screening process between 2014 and 2022. ...
+of these patients were also admitted to the ICU. ... patients were
+excluded as they were younger than 15 years and ... patients were
+excluded because they were dead on arrival. This leaves us with 167
+outcomes.
+
+This paragraph is not finished! Will finish after I start working on the
+real data. Will maybe add a table.
+
+## Statistical methods
+
+Bivariable logistic regression was used to determine unadjusted
+associations and multivariable logistic regression to determine adjusted
+associations between patient and process factors and OFI. Odds ratios
+(OR) with associated 95% confidence intervals was used with a 5%
+significance level.
+
+## Ethical considerations
+
+There are several ethical aspects to take into consideration regarding
+this study: voluntary participation, informed consent, anonymity,
+confidentiality, potential for harm and results communication.
+
+As described previously, all patients with trauma team activation or
+high ISS scores are included in the database used for this study. Having
+suffered severe trauma might not be something the patients or their
+relatives would want to be public information as the collected data
+contain for instance vital parameters and demise as well as type of
+trauma and time of arrival to the hospital, which might be sensitive
+information. Therefore, the ethical aspect of voluntary participation
+and informed consent has to be considered. The data in national
+databases is collected from all patients registered and is not based on
+consent, even though it might be informed, which is a drawback.
+
+To respect the anonymity and confidentiality aspect the statistical
+analysis was first conducted on synthetic data and later implemented on
+the data collected from the trauma registry and the trauma care quality
+database in order to protect personal numbers and their associated
+information. The anonymity might ease the discomfort among the involved
+patients and their relatives.
+
+There is no direct potential for harm by being included in the study as
+no intervention is made and it is an observational study.
+
+In regard to results communication, the results from research based on
+the data as well as a compilation of the original data is public. The
+data being accessible for research does not directly benefit the
+affected individual; however, it contributes greatly to the development
+of the hospital care in Sweden, making it more efficient and equal.
+Studies performed on data from the Karolinska University Hospital in
+Solna are made to improve local trauma care systems. This way, it can
+also benefit the society as improved local trauma care decreases trauma
+death rate and improves quality of care and thus patient safety.
+
+Based on these arguments it can be concluded that the advantages
+outweigh the disadvantages. The advantages being the contribution to the
+development of trauma systems which long-term improves patient safety
+and equality, and that the study protects the individual's anonymity and
+confidentiality by using synthetic data and later implementing it on the
+real data. The disadvantage being the potential emotional harm from the
+absence of voluntary participation.
+
+The study was approved by the Swedish Ethical Review Authority, approval
+numbers 2023-02975-02, 2021-03531, 2021-02541.
+
+# Results
+
+Table 1: deskriptiv tabell, stratifierad från utfallet, förslag är att
+använda:
+<https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html>
+
+Table 2: huvudresultat, förslag att använda:
+<https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html>
+presentera ojusterad (univariable logistisk regression) och justerad
+(multivariable log reg)
+
+Tillse korrekt presentation av resultaten i textformat:
+<https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html?q=tbl%20_%20re#inline_text_tbl_regression>
+
+Ytterligare tabeller, audit filtren? En flow chart för
+inklusion/exklusion?
+
+```         
 ```
 
-Abstract
-========
+## Participants
 
-Background
-----------
-Trauma is a leading cause of mortality and morbidity worldwide.
-A cornerstone in developing the trauma quality improvement programs is multidisciplinary mortality and morbidity review.
-The purpose of these is to identify opportunities for improvement (OFI), which are necessary in order to implement corrective actions.
-Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
-The aim of this study is to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
-
-Methods
--------
-This was a registry-based study using all patients between 2014 and 2022 from the Karolinska University Hospital in Solna who had been admitted to the ICU and who were reviewed regarding the presence of OFI by the mortality and morbidity conference.
-Bi-variable and multivariable logistic regression was used to assess the following patient and process factors and OFI among this group of patients: gender, age, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT), if the patient was intubated and Injury Severity Score (ISS).
-
-Results
--------
-
-Conclusion
-----------
-
-Abbreviations
-=============
-CT = computed tomography 
-GCS = Glasgow Coma Scale
-GOS = Glasgow Outcome Score
-ISS = Injury Severity Score 
-NISS = New Injury Severity Score 
-OFI = Opportunity For Improvement 
-RTS = Revised Trauma Score 
-TRISS = Trauma and Injury Severity Score
-
-Introduction
-============
-Trauma — the clinical entity of injury and the body's associated response — is a leading cause of mortality and morbidity worldwide.
-According to WHO, injury-related deaths constitute 4% of all deaths and take the lives of 4,4 million people yearly worldwide - more than HIV/AIDS, malaria, maternal mortality and tuberculosis combined.
-The top 3 of 5 causes of death in people aged 5-29 are injury-related; road traffic injuries, homicide and suicide.
-Each year, tens of millions of people suffer non-fatal injuries leading to treatment at hospitals in emergency departments and acute care visits.
-These injuries often lead to temporary or permanent disability and the need for medical and mental rehabilitation.
-Trauma exposure at a young age increases the risk for mental illness and suicide leading to increased risk of an unhealthy lifestyle associated with smoking, alcohol and substance abuse, chronic disease and cancer as well as societal problems such as crime, poverty and violence.
-The burden of disease from injuries vary depending on certain conditions; socioeconomic factors, gender, age and country.
-Men suffer from injuries more often than women.
-Across all age groups worldwide, the top three leading causes of injury among males are road traffic injuries, homicide and suicide; whereas the top three leading causes of injury among women are road traffic injuries, falls and suicide [1,2].
-
-According to the annual report from the Swedish Trauma registry (SweTrau) 2022, trauma was more common among men than women in Sweden — 65,3% among men compared to 34,7% among women.
-The most affected age group was the working population.
-The leading causes of trauma included; road traffic injuries, falls, weapons and stabs and blows.
-Previous years, road traffic injuries have been the leading cause of injury.
-However, since 2022 road traffic injuries and falls account for about the same proportion of mechanisms of injury.
-There are several variables that can be studied to assess primary patient care, for example time to first CT and acute interventions.
-Median time to first CT in the university hospitals of Sweden was 30 minutes in unconscious patients with GCS < 9.
-Acute interventions is defined as interventions performed immediately or within 24 hours after arrival to a hospital.
-The most common acute interventions in Swedish hospitals were: thoracic drains, wound revisions and fracture surgery followed by laparotomy with haemostasis, radiological interventions such as embolisation and stents, and thoracotomy [3].
-
-Albaaj et al.
-recently analyzed OFI in relation to different patient and process factors based on data from the Swedish trauma registry and data from the Karolinska University Hospital  in Solna.
-The study indicated that patients with moderate to severe trauma are at the highest odds of OFI [4].
-According to SweTrau 2022 41,1% of patients with New Injury Severity Score (NISS) > 15 who were at a university hospital were admitted to the ICU.
-The proportion of severely injured admitted to the ICU has decreased in recent years, however; patients with NISS > 15 still constitute a large proportion of the trauma patients treated in the ICU.
-A way to measure functional level after hospital discharge is Glasgow Outcome Scale (GOS).
-According to GOS, among patients with NISS > 15 17% were deceased, 20% suffered severe disability, 49% suffered moderate disability and 14% had good recovery in the university hospitals of Sweden [3].
-
-For all injuries, quality hospital care can reduce the amount of short- and long-term disability.
-Therefore, improving planning, access and organisation of trauma care systems including pre-hospital care, interhospital transportation, telecommunication and in-hospital care all play a central role in reducing overall impact after trauma [1].
-Disability-adjusted life-years (DALYs) is a way to measure post-discharge quality of life.
-Trauma accounts for one-tenth of DALYs, most of the burden in low and middle income countries (LMIC) [5, 6].
-Trauma costs approximately 3% of gross domestic product (GDP) yearly globally, as it mostly affects the working population [7].
-
-Trauma quality improvement programs
------------------------------------
-Many initiatives for quality improvement programs have been initiated in trauma and critical care based on the paradigm of Donabedian [8], stating that implementing proven structures and processes of care is the most effective way to improve outcomes.
-One example of a structure and process intervention in the intensive care unit is highly-staffed ICUs that has been associated with improved outcomes, as well as bundles to recognize and treat common conditions such as those for treating sepsis and ventilator-associated pneumonia [9].
-The intention of quality improvement is objective improvement in regards to preventability.
-A cornerstone in developing the trauma quality improvement programs is therefore multidisciplinary mortality and morbidity review.
-The purpose of these is to discuss preventable deaths and identifying opportunities for improvement (OFI) in trauma care regarding structure as well as clinical processes.
-This way, identifying OFIs will provide guidance as to what to focus specific efforts on.
-There are several other methods to improve trauma care.
-However, mortality and morbidity review has been shown to be a key method of addressing all necessary components of trauma improvement [10].
-
-Trauma audit filters are descriptions of specific actions that should be taken, time frames in which tests or treatments should be provided, or outcomes that are expected to occur in injured patients.
-Assessing patients whose care falls outside the frames of these audit filters and providing feedback to the clinicians involved in the patient's care provides means of correcting errors and improving future trauma care.
-The audit filters were introduced by the American College of Surgeons 1995; however, they have been criticized as they may not be directly linked to meaningful clinical or patient-centered outcomes [11].
-At the Karolinska University Hospital in Solna local audit filters are used, shown in Table 1.
-These local audit filters include: systolic blood pressure under 90, GCS < 9 and not intubated, ISS > 15 but not admitted to the ICU, ISS > 15 and no trauma team activation, no anticoagulation treatment within 72 h after traumatic brain, injury time to acute intervention more than 60 minutes, time to computed tomography (CT) more than 30 minutes, liver or spleen injury, cardiopulmonary resuscitation with thoracotomy, massive transfusion and death within 30 days after trauma.
-
-Injury score systems
---------------------
-When a patient first arrives at the hospital, injury system scores are used to assess how urgent the need of interventions to a patient is.
-There are several injury score systems: Injury Severity Score (ISS), New Injury Severity Score (NISS), Revised Trauma Score (RTS) and Trauma and Injury Severity Score (TRISS).
-
-ISS is an anatomical scoring system providing an overall score for patients with multiple injuries.
-Each injury is assigned an abbreviated injury scale (AIS) score and is allocated to one of six body regions.
-The three most severely injured body regions have their scores squared and added together to produce an ISS score between 0 and 75.
-If an injury is assigned an AIS of 6 the ISS score is automatically assigned to 75.
-Major trauma is considered when ISS > 15.
-The major drawback of ISS is that it only considers one injury for each body region.
-Therefore, if several severe injuries are in the same body region, they are not accounted for.
-NISS counters the problem of ISS being the sum of the square of the three most severe injuries, regardless of body region injured.
-NISS is therefore equal to or higher than ISS.
-
-RTS is a physiological scoring system scored from the data obtained in the first set of datas obtained on the patient and consists of Glasgow Coma Scale (GCS), respiratory rate and systolic blood pressure.
-RTS is heavily weighted towards GCS compared to respiratory rate and systolic blood pressure.
-It has been demonstrated to show accuracy at predicting death.
-
-TRISS has a maximum prediction in outcome when compared with the other scores.
-The score determines the probability of survival based on a formula consisting of ISS, RTS and the patient's age.
-TRISS is therefore both an anatomical and physiological scoring system [12].
-
-Preventable deaths in patients admitted to the ICU
---------------------------------------------------
-There are many studies on preventable deaths in trauma patients treated in the ICU both regarding the medical aspect as well as the human aspect.
-According to a study from Japan on the medical aspect, the primary medical causes of death after trauma are haemorrhage, sepsis/multiorgan failure and central nervous system injury.
-Delayed hemostatic procedures and transfusions were common areas for improvement in the hospital stage [13].
-The most prominent OFI in regards to haemorrhage was decision-making compared to errors in technical skill.
-The OFIs frequently involved the decision between radiology, surgery and further investigation [14].
-
-A study from The University of Pennsylvania looked at the human aspect of OFIs in the ICU.
-The ICU of this hospital uses telemedicine and telemonitoring to review critical patient events.
-The areas of improvement they identified were: team dynamics and communication, use of best practices, increased and standardized access to emergency resources and improvement of procedural technique.
-They recommend other ICUs with telemedicine capacity to adopt a similar strategy in order to present OFIs and educational discussions [15].
-Another study looked at how critical information was handled in the ICU and found that laboratory values and test results were the most frequently lost items.
-This points at the importance of communication and handling of critical information at handoffs of care [16].
-
-Preventability in studies is often defined as an event that would not have occurred had the patient received ordinary standards of care.
-Evaluating preventable deaths in patients admitted to the ICU can be a challenge as the patients have extensive medical history and it is harder to understand what the standard of care is for those patients.
-Furthermore, even though these patients receive the best of care, these patients still have a higher risk of developing complications or dying.
-Some might even be in such a bad condition that they die just when they have been admitted to the ICU from the ward or emergency care [9].
-
-Aim
----
-Many patients with severe trauma are admitted to the intensive care unit (ICU), but little is known about opportunities for improvement in this group of patients.
-This being due to the patients admitted to the ICU being more complex.
-Identifying OFIs in this group can therefore be more of a challenge.
-There is existing studies on different aspects of trauma care improvement, but studies focusing mainly on OFI in the ICU are few.
-The aim of this study is therefore to characterise OFI in adult trauma patients admitted to the ICU and assess how patient factors are associated with OFI in these patients.
-
-Methods 
-=======
-Study design 
-------------
-We conducted a retrospective single-centre cohort study based on data from the Karolinska University Hospital in Solna and from the Swedish Trauma registry.
-
-Setting
--------
-The Karolinska University Hospital in Solna is a level 1 trauma centre based on the definition by the American College of Surgeons (ACS) Committee on Trauma [17], which means that all patients in Stockholm with severe or suspected severe trauma are sent there.
-The hospital has direct access to radiology, surgery, emergency medicine, intensive care and consultants from relevant specialities [18].
-All patients are included in a mortality and morbidity screening combining individual review and specific audit filters by specialized nurses.
-The patients who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks.
-Examples of OFI include the need for better organization or more senior members taking part in the trauma care.
-The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.
-
-Participants
-------------
-The trauma registry includes all patients admitted with trauma team activation, regardless of Injury Severity Score (ISS), as well as patients admitted without trauma team activation but found to have an ISS of more than 9.
-We included all patients who had been included in the morbidity and mortality screening process between January 1, 2014 and June 1, 2022 and who were also admitted to the ICU.
-We excluded patients who were younger than 15 years and patients who were dead on arrival.
-
-Variables and data sources/measurements
----------------------------------------
-### Study outcome
-The study outcome will be on the presence of OFI in adult patients treated in the ICU based on patient and process factors.
-
-### Patient and process factors 
-The quantitative variables were factors chosen from the trauma registry, based on the locally used audit filters, standard epidemiological factors and factors registered in the Swedish National Trauma Registry.
-The categorical factors were sex, survival after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic blood pressure, working hours, weekend, time from arrival at the hospital until first computed tomography (CT) and if the patient was intubated.
-The continuous factors age and Injury Severity Score (ISS) were also included.
-Working hours was defined as between 8.00 a.m.
-and 5 p.m., or during a weekend, defined as Saturday or Sunday.
-
-Bias/ethical considerations
----------------------------
-We used R for all statistical analyses [19].
-A complete case analysis was conducted after handling missing values in systolic blood pressure, respiratory rate, and GCS score.
-We present sample characteristics using descriptive statistics.
-The statistical analysis was first conducted on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database in order to protect personal numbers and their associated information.
-
-Study size 
-----------
-In the data there are 167 outcomes. Will later explain how the study size was arrived at.
-
-Statistical methods
--------------------
-Bivariable logistic regression was used to determine unadjusted associations and multivariable logistic regression to determine adjusted associations between patient and process factors and OFI.
-Odds ratios (OR) with associated 95% confidence intervals was used with a 5% significance level.
-
-<!-- 
-(a) Describe all statistical methods, including those used to control
-for confounding
-(b) Describe any methods used to examine subgroups and interactions
-(c) Explain how missing data were addressed 
-(d) Cohort study: If applicable, explain how loss to follow-up was addressed 
-(e) Describe any sensitivity analyses
--->
-
-Results
-=======
-Table 1: deskriptiv tabell, stratifierad från utfallet, förslag är att använda: https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html
-
-Table 2: huvudresultat, förslag att använda: https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html 
-presentera ojusterad (univariable logistisk regression) och justerad (multivariable log reg)
-
-Tillse korrekt presentation av resultaten i textformat: https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html?q=tbl%20_%20re#inline_text_tbl_regression
-
-Ytterligare tabeller, audit filtren? En flow chart för inklusion/exklusion?
-
-Participants
-------------
-
+```{=html}
 <!-- (a) Report numbers of individuals at each stage of study—eg numbers
@@
 -->
+```
+## Descriptive data
 
-Descriptive data
-----------------
-
+```{=html}
 <!-- (a) Give characteristics of study participants (eg demographic,
@@
 -->
+```
+## Outcome data
 
-Outcome data
-------------
-
+```{=html}
 <!-- Cohort study — Report numbers of outcome events or summary measures
@@
 -->
+```
+## Main results
 
-Main results
-------------
-
+```{=html}
 <!-- (a) Give unadjusted estimates and, if applicable, confounder-adjusted
@@
 -->
+```
+## Other analyses
 
-Other analyses 
---------------
-
+```{=html}
 <!--Report other analyses done—eg analyses of subgroups and interactions,
@@
 -->
+```
+# Discussion
 
-```{r main, echo=FALSE}
-source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
-
-Discussion
-==========
-
+```{=html}
 <!--
@@
 -->
-
-Conclusion
-==========
+```
+# Conclusion
 
@@
 
-References
-==========
+# References
 

### 2024-02-21 main.R
commit 21/2

@@ -1,152 +1,389 @@
-## INTRUCTIONS 
-## This is your project's main script file and together with
-## manuscript.Rmd it provides and entry point for you and other people
-## coming to the project. The code in this file should give an outline
-## of the different steps conducted in your study, from importing data
-## to producing results.
+#INSTALLING PACKAGES, does not work with function count! 
+#install.packages("dplyr")
+library("dplyr")
+#install.packages("gtsummary")
+library("gtsummary")
 
-## This file should be relatively short, and most of the heavy
-## lifting should be done by specialised functions. These functions
-## live in the folder functions/ and you create a new function using
-## create_function().
+#TABLE1 Local audit filters
+auditfilter1 <- merged.data %>% 
+  select(VK_sap_less90, VK_gcs_less9_ej_intubTE, VK_iss_15_ej_iva, VK_iss_15_ej_TE, 
+         VK_ej_trombrof_TBI_72h, VK_mer_30min_DT, VK_leverskada, VK_mjaltskada, 
+         VK_hlr_thorak, VK_mass_transf, res_survival)
 
-## Source all functions (if you tick the box "Source on save" in
-## RStudio functions will be automatically sourced when you save
-## them). They all need to be sourced however when you compile your
-## manuscript file or run this file as a job, as that happens in a
-## clean R session.
+auditfilter1$VK_sap_less90 <- ifelse(auditfilter1$VK_sap_less90 == "Ja", "Yes", 
+                                     ifelse(auditfilter1$VK_sap_less90 == "nej", NA,
+                                            ifelse(auditfilter1$VK_sap_less90 == "Nej", NA, NA)))
 
+auditfilter1$VK_gcs_less9_ej_intubTE <- ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Ja", "Yes",
+                                               ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "nej", NA,
+                                                      ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Nej", NA, NA)))                            
 
+auditfilter1$VK_iss_15_ej_iva <- ifelse(auditfilter1$VK_iss_15_ej_iva == "Ja", "Yes",
[…]

### 2024-02-21 manuscript.Rmd

@@
 post-discharge quality of life. Trauma accounts for one-tenth of DALYs,
-most of the burden in low and middle income countries (LMIC) [@5, @6].
+most of the burden in low- and middle-income countries (LMIC) [@5, @6].
 Trauma costs approximately 3% of gross domestic product (GDP) yearly
@@
 and ventilator-associated pneumonia [@9]. The intention of quality
-improvement is objective improvement in regards to preventability. A
+improvement is objective improvement in regard to preventability. A
 cornerstone in developing the trauma quality improvement programs is
@@
 hemostatic procedures and transfusions were common areas for improvement
-in the hospital stage [@13]. The most prominent OFI in regards to
+in the hospital stage [@13]. The most prominent OFI in regard to
 haemorrhage was decision-making compared to errors in technical skill.
@@
 
-## Aim
-
 Many patients with severe trauma are admitted to the intensive care unit
@@
 trauma care improvement, but studies focusing mainly on OFI in the ICU
-are few. The aim of this study is therefore to characterise OFI in adult
-trauma patients admitted to the ICU and assess how patient factors are
-associated with OFI in these patients.
+are few.
+
+## Aim
+
+The aim of this study is to characterise OFI in adult trauma patients
+admitted to the ICU and assess how patient factors are associated with
+OFI in these patients.
 
@@
 
-This paragraph is not finished! Will finish after I start working on the
-real data. Will maybe add a table.
-
 ## Statistical methods
@@
 
-Table 1: deskriptiv tabell, stratifierad från utfallet, förslag är att
-använda:
-<https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html>
+[n] patients were included in the trauma registry and trauma care
+quality database between 2017 and 2022. [n] patients were excluded,
+leaving a total of [n] patients eligible for the study. [n] were
+excluded because they were under 15 years old. [n] patients were
+excluded because they were dead on arrival. [n] were excluded because
+they were not treated in the ICU, and [n] patients were excluded because
+they had missing data (?).
 
-Table 2: huvudresultat, förslag att använda:
-<https://www.danieldsjoberg.com/gtsummary/reference/tbl_regression.html>
-presentera ojusterad (univariable logistisk regression) och justerad
-(multivariable log reg)
+[n{%}] unique patients of the [n] patients included were flagged with
+OFI. Table 1 illustrates the number of patients flagged per audit filter
+used by the Karolinska University Hospital in Solna.
 
-Tillse korrekt presentation av resultaten i textformat:
-<https://www.danieldsjoberg.com/gtsummary/articles/inline_text.html?q=tbl%20_%20re#inline_text_tbl_regression>
+Table 2 present sample characteristics. [n{%}] patients were male and
+[n{%}] patients were female. [n{%}] patients had no intubation. Most
+patients of the intubated ones were intubated for xx days [n{%}]. The
+number of patients who arrived outside working hours - on Saturday or
+Sunday or after 5 pm or before 8 am were [n{%}]. [n{%}] patients had an
+ASA-score of 1-2 before arrival to the hospital, and [n{%}] patients had
+an ASA-score of 3-7 before arrival to the hospital. [n{%}] patients died
+within 30 days after hospitalization.
 
-Ytterligare tabeller, audit filtren? En flow chart för
-inklusion/exklusion?
+Table 3 shows the unadjusted and adjusted associations of the selected
+patient and process factors in relation to OFI. In the unadjusted
+analysis the factors that were significantly associated with the
+presence of OFI were: ... Male
+`r inline_text(table3a, variable = Sex, level = "Male")` Female
+`r inline_text(table3a, variable = Sex, level = "Female")` Age
+`r inline_text(table3a, variable = Age)` No intubation
+`r inline_text(table3a, variable = Intubation, level = "No intubation")`
+Respiratory rate `r inline_text(table3a, variable = RespiratoryRate)`
+Systolic blood pressure
+`r inline_text(table3a, variable = SystolicBloodPressure)` Glasgow Coma
+Scale `r inline_text(table3a, variable = GlasgowComaScale)` Jour
+`r inline_text(table3a, variable = Jour, level = "Yes")` Time to first
+CT `r inline_text(table3a, variable = TimeFCT)` Time to first
+intervention `r inline_text(table3a, variable = TimeFInt)` ASA preinjury
+1-2 `r inline_text(table3a, variable = ASApreinjury, level = "ASA 1-2")`
+Death within 30 days
+`r inline_text(table3a, variable = Survival, level = "Dead")`
 
-```         
-```
+In the adjusted analysis the following factors demonstrated significant
+association with OFI: ...
+
+x was significantly associated with increased odds of OFI in both the
+unadjusted and adjusted analyses.
+
+x were not significantly associated with OFI in the unadjusted or the
+adjusted analyses.
 

### 2024-02-24 manuscript.Rmd
commit 24/2

@@
 linked to meaningful clinical or patient-centered outcomes [@11]. At the
-Karolinska University Hospital in Solna local audit filters are used,
-shown in Table 1. These local audit filters include: systolic blood
-pressure under 90, GCS \< 9 and not intubated, ISS \> 15 but not
-admitted to the ICU, ISS \> 15 and no trauma team activation, no
-anticoagulation treatment within 72 h after traumatic brain injury, time
-to acute intervention more than 60 minutes, time to CT more than 30
-minutes, liver or spleen injury, cardiopulmonary resuscitation with
-thoracotomy, massive transfusion and death within 30 days after trauma.
+Karolinska University Hospital in Solna local audit filters are used.
+These local audit filters include: systolic blood pressure under 90, GCS
+\< 9 and not intubated, ISS \> 15 but not admitted to the ICU, ISS \> 15
+and no trauma team activation, no anticoagulation treatment within 72 h
+after traumatic brain injury, time to acute intervention more than 60
+minutes, time to CT more than 30 minutes, liver or spleen injury,
+cardiopulmonary resuscitation with thoracotomy, massive transfusion and
+death within 30 days after trauma.
 

### 2024-02-24 functions/example_function.R

@@ -1,7 +1,390 @@
-## This file shows how to write a function in R
+#INSTALLING PACKAGES, does not work with function count! 
+#install.packages("dplyr")
+library("dplyr")
+#install.packages("gtsummary")
+library("gtsummary")
 
-example_function <- function() {
-    ## This is a comment and will not be interpreted by R
-    x <- 1 + 2
-    return (x)
+#TABLE1 Local audit filters
+#auditfilter1 <- merged.data %>% 
+#  select(VK_sap_less90, VK_gcs_less9_ej_intubTE, VK_iss_15_ej_iva, VK_iss_15_ej_TE, 
+#         VK_ej_trombrof_TBI_72h, VK_mer_30min_DT, VK_leverskada, VK_mjaltskada, 
+#         VK_hlr_thorak, VK_mass_transf, res_survival)
+
+#auditfilter1$VK_sap_less90 <- ifelse(auditfilter1$VK_sap_less90 == "Ja", "Yes", 
+#                                    ifelse(auditfilter1$VK_sap_less90 == "nej", NA,
+#                                           ifelse(auditfilter1$VK_sap_less90 == "Nej", NA, NA)))
+                                    
+#auditfilter1$VK_gcs_less9_ej_intubTE <- ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Ja", "Yes",
+#                                              ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "nej", NA,
+#                                                     ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Nej", NA, NA)))                            
+
+#auditfilter1$VK_iss_15_ej_iva <- ifelse(auditfilter1$VK_iss_15_ej_iva == "Ja", "Yes",
+#                                               ifelse(auditfilter1$VK_iss_15_ej_iva == "nej", NA,
+#                                                      ifelse(auditfilter1$VK_iss_15_ej_iva == "Nej", NA, NA)))     
+
+#auditfilter1$VK_iss_15_ej_TE <- ifelse(auditfilter1$VK_iss_15_ej_TE == "Ja", "Yes",
+#                                        ifelse(auditfilter1$VK_iss_15_ej_TE == "nej", NA,
+#                                               ifelse(auditfilter1$VK_iss_15_ej_TE == "Nej", NA, 
+#                                                      ifelse(auditfilter1$VK_iss_15_ej_TE == "nn", NA,NA))))     
+
[…]

### 2024-02-27 manuscript.Rmd
27/2

@@
 logistic regression was used to assess the following patient and process
-factors and OFI among this group of patients: gender, age, survival
-after 30 days, Glasgow Coma Scale (GCS), respiratory rate, systolic
-blood pressure, working hours, weekend, time from arrival at the
-hospital until first computed tomography (CT), if the patient was
-intubated and Injury Severity Score (ISS).
+factors and OFI among this group of patients: gender, age, if the
+patient was intubated or not and how many days, respiratory rate,
+systolic blood pressure, Glasgow Coma Scale (GCS), Injury Severity Score
+(ISS), on call, time from arrival at the hospital until first computed
+tomography (CT), ASA preinjury and survival after 30 days.
 
@@
 from the Karolinska University Hospital in Solna and from the Swedish
-Trauma registry. We used R for all statistical analyses [@19]. A
-complete case analysis was conducted after handling missing values in
-systolic blood pressure, respiratory rate and GCS score. We present
-sample characteristics using descriptive statistics.
+Trauma registry. We used R for all statistical analyses [@19]. For
+patients where hospital values for systolic blood pressure, respiratory
+rate or GCS score were missing, prehospital values were used instead. A
+complete case analysis was conducted after handling missing values. We
+present sample characteristics using descriptive statistics.
 
@@
 the morbidity and mortality screening process between January 1, 2014
-and June 1, 2022 and who were also admitted to the ICU. We excluded
+and February 1, 2023 and who were also admitted to the ICU. We excluded
 patients who were younger than 15 years and patients who were dead on
@@
 factors and factors registered in the Swedish Trauma Registry. The
-categorical factors were sex, survival after 30 days, GCS, respiratory
-rate, systolic blood pressure, working hours, weekend, time from arrival
-at the hospital until first CT, if the patient was intubated, age and
-Injury Severity Score (ISS). Working hours was defined as between 8.00
-a.m. and 5 p.m., or during a weekend, defined as Saturday or Sunday.
+categorical factors were sex, age, if the patient was intubated or not
+and how many days, respiratory rate, systolic blood pressure, Glasgow
+Coma Scale (GCS), Injury Severity Score (ISS), on call, time from
+arrival at the hospital until first computed tomography (CT), ASA
+preinjury and survival after 30 days. On duty was defined as before 8.00
+a.m., after 5 p.m., or during a weekend - defined as Saturday or Sunday.
 
@@
 
-There were ... patients in the trauma registry. ... with trauma team
-activation regardless of ISS score, ... patients admitted without trauma
-team activation but with ISS score \> 9. .... patients were included in
-the morbidity and mortality screening process between 2014 and 2022. ...
-of these patients were also admitted to the ICU. ... patients were
-excluded as they were younger than 15 years and ... patients were
-excluded because they were dead on arrival. This leaves us with 167
-outcomes.
+There were ... patients in the trauma registry between January 1 2014
+and February 1 2023. ... with trauma team activation regardless of ISS
+score, ... patients admitted without trauma team activation but with ISS
+score \> 9. .... patients were included in the morbidity and mortality
+screening process between 2014 and 2022. ... of these patients were also
+admitted to the ICU. ... patients were excluded as they were younger
+than 15 years and ... patients were excluded because they were dead on
+arrival. This leaves us with 167 outcomes.
 
@@
 
-[n] patients were included in the trauma registry and trauma care
-quality database between 2017 and 2022. [n] patients were excluded,
-leaving a total of [n] patients eligible for the study. [n] were
-excluded because they were under 15 years old. [n] patients were
-excluded because they were dead on arrival. [n] were excluded because
-they were not treated in the ICU, and [n] patients were excluded because
-they had missing data (?).
+14,022 patients were included in the trauma registry and trauma care
+quality database between 2017 and 2023. 12,278 patients were excluded,
+leaving a total of 1,744 patients eligible for the study. 11,343 were
+excluded because they were not treated in the ICU. 3 patients were
+excluded because they were under 15 years old. 6 patients were excluded
+because they were dead on arrival, and 926 patients were excluded
+because there was no data on the presence or absence OFI.
 
-[n{%}] unique patients of the [n] patients included were flagged with
-OFI. Table 1 illustrates the number of patients flagged per audit filter
-used by the Karolinska University Hospital in Solna.
+[n] patients were excluded because there was missing data, leaving a
+total of [n] patients included in the study.
 
-Table 2 present sample characteristics. [n{%}] patients were male and
+Table 1 present sample characteristics. [n{%}] patients were male and
 [n{%}] patients were female. [n{%}] patients had no intubation. Most
@@
 
-Table 3 shows the unadjusted and adjusted associations of the selected
+Table 2 shows the unadjusted and adjusted associations of the selected
 patient and process factors in relation to OFI. In the unadjusted
@@
 
-```{=html}
-<!-- (a) Report numbers of individuals at each stage of study—eg numbers
-potentially eligible, examined for eligibility, confirmed eligible,
-included in the study, completing follow-up, and analysed
+<!-- Flowchart -->
 
-(b) Give reasons for non-participation at each stage
-
-(c) Consider use of a flow diagram
-
--->
-```
 ## Descriptive data
 
-```{=html}
-<!-- (a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
+<!-- Sample characteristics - Table1, do I need to display number of missing for each variable? -->
 
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
--->
-```
-## Outcome data
-
-```{=html}
-<!-- Cohort study — Report numbers of outcome events or summary measures
-over time
--->
-```
 ## Main results
 
-```{=html}
-<!-- (a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
+<!--  Adjusted and unadjusted analysis - Table 2 -->
 
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
--->
-```
-## Other analyses
-
-```{=html}
-<!--Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses 
--->
-```
 # Discussion
 
-```{=html}
-<!--
+### Key results
 
-Key results
------------
-Summarise key results with reference to study objectives
+We found that --- was associated with higher odds of OFI. Previous
+research ...
 
-Limitations
------------
+These findings could be because ...
 
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias
+\-- was associated with OFI in the unadjusted analysis but not in the
+adjusted analysis.
 
-Interpretation
---------------
+1.  add sources from previous studies
 
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from
-similar studies, and other relevant evidence
+2.  studies to back up the results from the analyses
 
-Generalisability
-----------------
+### Limitations
 
-Discuss the generalisability (external validity) of the study results
+There were several limitations in our study. The selection for review
+for the morbidity and mortality review relied mostly on the local audit
+filters used at the Karolinska Institute Hospital in Solna. This leads
+to a risk of misplacement where some patients with OFI were not selected
+by for the review based on the audit filters although there was an OFI.
 
--->
-```
+Another limtation is that some relevant factors, such as ----, were not
+part of the Swedish Trauma registry and therefore not adjusted for in
+the analysis.
+
+The study was a single-center study and illustrate the situation in
+Stockholm. More study is needed on the area of OFI in patients admitted
+to the ICU. A suggestion would be to conduct a multi-center study on
+this group of patients to further investigate areas for improvement and
+to further develop trauma quality programs.
+
+Future studies could also focus on ---- which was seen to be associated
+with OFI in these patients.
+
+<!--Lägg till 2 st förslag på explicita fortsättningsstudier! -->
+
+### Generalisability
+
+The study is conducted on --- outcomes. There were few outcomes because
+we included only those who were admitted to the ICU, \> 14 years old,
+not dead on arrival and had a registration for the presence or absence
+of OFI. The external validity is therefore compromised due to the
+limited size of the Swedish database. However, ...
+
 # Conclusion
+
+---- among patients admitted to the ICU have higher odds of OFI. These
+factors can be seen as new targets for further reseach to further
+supplement trauma quality improvement programs.
 

### 2024-02-27 functions/example_function.R

@@ -1,94 +1,21 @@
-#INSTALLING PACKAGES, does not work with function count! 
+#INSTALLING PACKAGES
+#devtools::install_github("martingerdin/noacsr")
+#devtools::install_github("martingerdin/rofi")
+library(dotenv)
+library(noacsr)
+library(rofi)
+noacsr::source_all_functions()
+data <- import_data()
+
+merged.data <- merge_data(data)
+merged.data$ofi <- create_ofi(merged.data)
+
+
 #install.packages("dplyr")
-library("dplyr")
+library(dplyr)
 #install.packages("gtsummary")
-library("gtsummary")
+library(gtsummary)
 
-#TABLE1 Local audit filters
-#auditfilter1 <- merged.data %>% 
-#  select(VK_sap_less90, VK_gcs_less9_ej_intubTE, VK_iss_15_ej_iva, VK_iss_15_ej_TE, 
-#         VK_ej_trombrof_TBI_72h, VK_mer_30min_DT, VK_leverskada, VK_mjaltskada, 
-#         VK_hlr_thorak, VK_mass_transf, res_survival)
-
-#auditfilter1$VK_sap_less90 <- ifelse(auditfilter1$VK_sap_less90 == "Ja", "Yes", 
-#                                    ifelse(auditfilter1$VK_sap_less90 == "nej", NA,
-#                                           ifelse(auditfilter1$VK_sap_less90 == "Nej", NA, NA)))
-                                    
-#auditfilter1$VK_gcs_less9_ej_intubTE <- ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Ja", "Yes",
-#                                              ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "nej", NA,
-#                                                     ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Nej", NA, NA)))                            
-
-#auditfilter1$VK_iss_15_ej_iva <- ifelse(auditfilter1$VK_iss_15_ej_iva == "Ja", "Yes",
-#                                               ifelse(auditfilter1$VK_iss_15_ej_iva == "nej", NA,
-#                                                      ifelse(auditfilter1$VK_iss_15_ej_iva == "Nej", NA, NA)))     
-
-#auditfilter1$VK_iss_15_ej_TE <- ifelse(auditfilter1$VK_iss_15_ej_TE == "Ja", "Yes",
[…]

### 2024-02-28 functions/example_function.R
28/2

@@ -94,9 +94,8 @@ excluded1
 ##CLEANING DATA
 subdat <- merged.data %>%
   select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
-         res_survival, pre_intubated, ed_intubated, iva_vardtillfallen_n, dt_ed_first_ct, ISS,
-         Deceased, DeceasedDate, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
-         host_care_level, hosp_vent_days, dt_ed_emerg_proc, pt_asa_preinjury, iva_dagar_n, pre_gcs_sum, 
+         res_survival, pre_intubated, ed_intubated, dt_ed_first_ct, ISS, DateTime_ArrivalAtHospital, FirstTraumaDT_NotDone,
+         host_care_level, hosp_vent_days, pt_asa_preinjury, pre_gcs_sum, 
          pre_rr_value, pre_sbp_value, Fr1.12, ed_rr_rtscat, ed_sbp_rtscat, pre_rr_rtscat, pre_sbp_rtscat)
 
 #Converting subdat$ofi to logical so subset can be used 
@@ -111,15 +110,15 @@ adult <- subset(iva, subset = (pt_age_yrs > 14))
 #Deceased on arrival 
 alive <- subset(adult, subset = (Fr1.12 == 2 | is.na(Fr1.12)))
 
-#Only showing the subset with TRUE for ofi in patients over 14 yrs old                
-ofi <- subset(iva, subset = (ofi == TRUE | ofi == FALSE))
+#Removing ofi = NA            
+ofi <- alive %>% subset(!is.na(ofi))
 
 
 #DEFINING VARIABLES FOR TABLE 1 
 #Gender
 ofi$Sex <- ifelse(ofi$pt_Gender == 1, "Male", 
                   ifelse(ofi$pt_Gender == 2, "Female", 
-                         ifelse(ofi$pt_Gender == 999, NA,NA)))
+                         ifelse(ofi$pt_Gender == 999, NA, NA)))
 
 #Age
 ofi$Age <- ofi$pt_age_yrs
@@ -150,11 +149,11 @@ ofi$rts_rr <- ifelse(ofi$RespiratoryRate == 0, 0,
                                     ifelse(ofi$RespiratoryRate %in% 10:29, 3,
                                            ifelse(ofi$RespiratoryRate > 29, 4, NA)))))
 
[…]

### 2024-03-09 manuscript.Rmd
9/3

@@
 review and specific audit filters by specialized nurses. The patients
-who are identified as having higher probability for OFI are discussed in
-a multidisciplinary conference held every 6-8 weeks. Examples of OFI
+who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI
 include the need for better organization or more senior members taking
-part in the trauma care. The presence or absence of OFI is decided among
-the participants of the conference and then recorded in the trauma care
-quality database.
+part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.
 
@@
 
-There were ... patients in the trauma registry between January 1 2014
-and February 1 2023. ... with trauma team activation regardless of ISS
-score, ... patients admitted without trauma team activation but with ISS
-score \> 9. .... patients were included in the morbidity and mortality
-screening process between 2014 and 2022. ... of these patients were also
-admitted to the ICU. ... patients were excluded as they were younger
-than 15 years and ... patients were excluded because they were dead on
-arrival. This leaves us with 167 outcomes.
+There were 14,022 patients in the trauma registry between January 1 2014
+and February 1 2023. 2679 of these patients were 
+admitted to the ICU. 3 patients were then excluded as they were younger
+than 15 years and 6 patients were excluded because they were dead on
+arrival. 928 patients were excluded because there was no registration on whether there was an OFI or not. 293 patients were excluded because they had missing values in one or more variables.
+This leaves us with 143 outcomes for OFI, and 1,306 outcomes for "No OFI".
+
+#Detta blir upprepning då det redan står längre ner i resultatdelen?
 
@@
 As described previously, all patients with trauma team activation or
-high ISS scores are included in the database used for this study. Having
-suffered severe trauma might not be something the patients or their
-relatives would want to be public information as the collected data
-contain for instance vital parameters and demise as well as type of
-trauma and time of arrival to the hospital, which might be sensitive
+high ISS scores are included in the database used for this study.
+Having suffered severe trauma might not be something the patients or their relatives would want to be public information as the collected data contain for instance vital parameters and demise as well as type of trauma and time of arrival to the hospital, which might be sensitive
 information. Therefore, the ethical aspect of voluntary participation
@@
 # Results
+```{r}
+library(gtsummary)
+source("main.R")
+```
 
@@
 quality database between 2017 and 2023. 12,278 patients were excluded,
-leaving a total of 1,744 patients eligible for the study. 11,343 were
+leaving a total of 1,742 patients eligible for the study. 11,343 were
 excluded because they were not treated in the ICU. 3 patients were
 excluded because they were under 15 years old. 6 patients were excluded
-because they were dead on arrival, and 926 patients were excluded
+because they were dead on arrival, and 928 patients were excluded
 because there was no data on the presence or absence OFI.
 
-[n] patients were excluded because there was missing data, leaving a
-total of [n] patients included in the study.
+293 patients were excluded because there was missing data, leaving a
+total of 1,449 patients included in the study.
 
-Table 1 present sample characteristics. [n{%}] patients were male and
-[n{%}] patients were female. [n{%}] patients had no intubation. Most
-patients of the intubated ones were intubated for xx days [n{%}]. The
-number of patients who arrived outside working hours - on Saturday or
-Sunday or after 5 pm or before 8 am were [n{%}]. [n{%}] patients had an
-ASA-score of 1-2 before arrival to the hospital, and [n{%}] patients had
-an ASA-score of 3-7 before arrival to the hospital. [n{%}] patients died
-within 30 days after hospitalization.
+Table 1 present sample characteristics.
+1,096 (76%) patients were male and 353 (24%) patients were female.
+846 (58%) patients had no intubation, 241 (17%) patients were intubated for 1-7 days and 113 (7,8%) patients were intubated for more than 7 days.
+The number of patients who arrived outside working hours — on Saturday or Sunday or after 5 pm or before 8 am were 1,042 (72%).
+1,064 (73%) patients were equal to or less than 7 days in the ICU, and 385 (27%) patients were more than 7 days in the ICU.
+1,080 (75%) patients had an ASA-score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA-score of 3-7 before arrival to the hospital.
+286 (20%) patients died within 30 days after hospitalization.
 
@@
 analysis the factors that were significantly associated with the
-presence of OFI were: ... Male
-`r inline_text(table3a, variable = Sex, level = "Male")` Female
-`r inline_text(table3a, variable = Sex, level = "Female")` Age
-`r inline_text(table3a, variable = Age)` No intubation
-`r inline_text(table3a, variable = Intubation, level = "No intubation")`
-Respiratory rate `r inline_text(table3a, variable = RespiratoryRate)`
-Systolic blood pressure
-`r inline_text(table3a, variable = SystolicBloodPressure)` Glasgow Coma
-Scale `r inline_text(table3a, variable = GlasgowComaScale)` Jour
-`r inline_text(table3a, variable = Jour, level = "Yes")` Time to first
-CT `r inline_text(table3a, variable = TimeFCT)` Time to first
-intervention `r inline_text(table3a, variable = TimeFInt)` ASA preinjury
-1-2 `r inline_text(table3a, variable = ASApreinjury, level = "ASA 1-2")`
-Death within 30 days
-`r inline_text(table3a, variable = Survival, level = "Dead")`
+presence of OFI were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`,
+≤ 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "≤ 7 days")` and death within 30 days after hospitalization `r inline_text(table3a, variable = Survival, level = "Dead")`.
 
 In the adjusted analysis the following factors demonstrated significant
-association with OFI: ...
+association with OFI: Revised Trauma Score `r inline_text(table3b, variable = RTS)` and ≤ 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "≤ 7 days")`.
 
-x was significantly associated with increased odds of OFI in both the
+RTS and ≤ 7 days in the ICU were significantly associated with increased odds of OFI in both the
 unadjusted and adjusted analyses.
 
-x were not significantly associated with OFI in the unadjusted or the
-adjusted analyses.
+Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury were not significantly associated with OFI in the unadjusted or the adjusted analyses.
+Female sex, intubation > 7 days, > 7 days in the ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could not be analysed with logistic regression.
 
@@
 
-We found that --- was associated with higher odds of OFI. Previous
-research ...
+We found that RTS and ≤ 7 days in the ICU were associated with higher odds of OFI. Previous
+research has shown that ...
 
@@
 
-\-- was associated with OFI in the unadjusted analysis but not in the
+Death within 30 days after hospitalisation was associated with OFI in the unadjusted analysis but not in the
 adjusted analysis.
@@
 
-The study is conducted on --- outcomes. There were few outcomes because
+The study is conducted on 143 outcomes of OFI. There were few outcomes because
 we included only those who were admitted to the ICU, \> 14 years old,

### 2024-03-09 main.R

@@ -1,94 +1,21 @@
-#INSTALLING PACKAGES, does not work with function count! 
+#INSTALLING PACKAGES
+#devtools::install_github("martingerdin/noacsr")
+#devtools::install_github("martingerdin/rofi")
+library(dotenv)
+library(noacsr)
+library(rofi)
+noacsr::source_all_functions()
+data <- import_data()
+
+merged.data <- merge_data(data)
+merged.data$ofi <- create_ofi(merged.data)
+
+
 #install.packages("dplyr")
-library("dplyr")
+library(dplyr)
 #install.packages("gtsummary")
-library("gtsummary")
+library(gtsummary)
 
-#TABLE1 Local audit filters
-auditfilter1 <- merged.data %>% 
-  select(VK_sap_less90, VK_gcs_less9_ej_intubTE, VK_iss_15_ej_iva, VK_iss_15_ej_TE, 
-         VK_ej_trombrof_TBI_72h, VK_mer_30min_DT, VK_leverskada, VK_mjaltskada, 
-         VK_hlr_thorak, VK_mass_transf, res_survival)
-
-auditfilter1$VK_sap_less90 <- ifelse(auditfilter1$VK_sap_less90 == "Ja", "Yes", 
-                                     ifelse(auditfilter1$VK_sap_less90 == "nej", NA,
-                                            ifelse(auditfilter1$VK_sap_less90 == "Nej", NA, NA)))
-
-auditfilter1$VK_gcs_less9_ej_intubTE <- ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Ja", "Yes",
-                                               ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "nej", NA,
-                                                      ifelse(auditfilter1$VK_gcs_less9_ej_intubTE == "Nej", NA, NA)))                            
-
-auditfilter1$VK_iss_15_ej_iva <- ifelse(auditfilter1$VK_iss_15_ej_iva == "Ja", "Yes",
-                                        ifelse(auditfilter1$VK_iss_15_ej_iva == "nej", NA,
-                                               ifelse(auditfilter1$VK_iss_15_ej_iva == "Nej", NA, NA)))     
-
-auditfilter1$VK_iss_15_ej_TE <- ifelse(auditfilter1$VK_iss_15_ej_TE == "Ja", "Yes",
-                                       ifelse(auditfilter1$VK_iss_15_ej_TE == "nej", NA,
[…]

### 2024-03-11 manuscript.Rmd
11/3

@@
 
-There were 14,022 patients in the trauma registry between January 1 2014
-and February 1 2023. 2679 of these patients were 
+There were 14,022 patients in the trauma registry between January 1 2013
+and February 28 2023. 2679 of these patients were 
 admitted to the ICU. 3 patients were then excluded as they were younger
@@
 # Results
+## Participants
+```{r, include=FALSE}
+source("functions/flowchart.R")
+```
 ```{r}
-library(gtsummary)
-source("main.R")
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+
+connectGrob(vert$eligible, excluded, type = "L")
+connectGrob(vert$included, excluded1, type = "L")
+
+vert
+excluded
+excluded1
+```
+
+## Descriptive data
+```{r include_scripts, echo=FALSE, include=FALSE}
+source("functions/table1.R")
+```
+```{r}
+table2
+```
+
+## Main results
+```{r, include=FALSE}
+source("functions/table2.R")
+```
+```{r}
+table3_merge
 ```
@@
 Female sex, intubation > 7 days, > 7 days in the ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could not be analysed with logistic regression.
-
-## Participants
-
-<!-- Flowchart -->
-
-## Descriptive data
-
-<!-- Sample characteristics - Table1, do I need to display number of missing for each variable? -->
-
-## Main results
-
-<!--  Adjusted and unadjusted analysis - Table 2 -->
 

### 2024-03-11 functions/flowchart.R

@@ -1,0 +1,71 @@
+
+#FLOWCHART: included/excluded
+#install.packages("Gmisc")
+library(Gmisc, quietly = TRUE)
+library(glue)
+#install.packages("htmlTable")
+library(htmlTable)
+library(grid)
+library(magrittr)
+
+org_cohort <- boxGrob(glue("Total patient cases in trauma quality database",
+                           "n = {pop}",
+                           pop = txtInt(14022),
+                           .sep = "\n"))
+eligible <- boxGrob(glue("Eligible",
+                         "n = {pop}",
+                         pop = txtInt(1742),
+                         .sep = "\n"))
+included <- boxGrob(glue("Included (n = {incl}):",
+                         "- OFI: {ofi1}",
+                         "- No OFI: {ofi2}",
+                         ofi1 = 143,
+                         ofi2 = 1306,
+                         incl = txtInt(1449),
+                         .sep = "\n"),
+                    just = "left")
+excluded <- boxGrob(glue("Excluded (n = {tot}):",
+                         " - Not admitted to the ICU: {icu}",
+                         " - Patients < 15 years: {age}",
+                         " - Dead on arrival: {doa}",
+                         " - No data on OFI: {ofi}",
+                         tot = 12278,
+                         icu = 14022-2679,
+                         age = 2679-2676,
+                         doa = 2676-2670,
+                         ofi = 2670-1742,
+                         .sep = "\n"),
+                    just = "left")
+excluded1 <- boxGrob(glue("Excluded: missing data (n = {x})",
+                          x = 1742 - 1449,
+                          .sep = "\n"),
+                     just = "left")
+
+grid.newpage()
+vert <- spreadVertical(org_cohort,
+                       eligible = eligible,
+                       included = included)
+
+# Move excluded box
+excluded <- moveBox(excluded,
+                    x = 0.8,
+                    y = 0.7)
+
+excluded1 <- moveBox(excluded1,
+                     x = 0.8,
+                     y = 0.4)
+
+# Connect boxes vertically
+for (i in 1:(length(vert) - 1)) {
+  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
+    print
+}
+
+# Connect excluded box horizontally
+connectGrob(vert$eligible, excluded, type = "L")
+connectGrob(vert$included, excluded1, type = "L")
+
+# Print boxes
+vert
+excluded
+excluded1

### 2024-03-11 functions/table1.R

@@ -27,69 +27,6 @@ library(htmlTable)
 library(grid)
 library(magrittr)
 
-org_cohort <- boxGrob(glue("Total patient cases in trauma quality database",
-                           "n = {pop}",
-                           pop = txtInt(14022),
-                           .sep = "\n"))
-eligible <- boxGrob(glue("Eligible",
-                         "n = {pop}",
-                         pop = txtInt(1742),
-                         .sep = "\n"))
-included <- boxGrob(glue("Included (n = {incl}):",
-                          "- OFI: {ofi1}",
-                          "- No OFI: {ofi2}",
-                          ofi1 = 143,
-                          ofi2 = 1306,
-                         incl = txtInt(1449),
-                         .sep = "\n"),
-                    just = "left")
-excluded <- boxGrob(glue("Excluded (n = {tot}):",
-                         " - Not admitted to the ICU: {icu}",
-                         " - Patients < 15 years: {age}",
-                         " - Dead on arrival: {doa}",
-                         " - No data on OFI: {ofi}",
-                         tot = 12278,
-                         icu = 14022-2679,
-                         age = 2679-2676,
-                         doa = 2676-2670,
-                         ofi = 2670-1742,
-                         .sep = "\n"),
-                    just = "left")
-excluded1 <- boxGrob(glue("Excluded: missing data (n = {x})",
-                          x = 1742 - 1449,
-                          .sep = "\n"),
-                     just = "left")
-
-grid.newpage()
-vert <- spreadVertical(org_cohort,
-                       eligible = eligible,
-                       included = included)
-
-# Move excluded box
-excluded <- moveBox(excluded,
-                    x = 0.8,
-                    y = 0.7)
-
-excluded1 <- moveBox(excluded1,
-                     x = 0.8,
-                     y = 0.4)
-
-# Connect boxes vertically
-for (i in 1:(length(vert) - 1)) {
-  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
-    print
-}
-
-# Connect excluded box horizontally
-connectGrob(vert$eligible, excluded, type = "L")
-connectGrob(vert$included, excluded1, type = "L")
-
-# Print boxes
-vert
-excluded
-excluded1
-
-
 ##CLEANING DATA
 subdat <- merged.data %>%
   select(ofi, pt_Gender, pt_age_yrs,  ed_gcs_sum, ed_sbp_value, ed_rr_value, 
[…]

### 2024-03-11 functions/table2.R

@@ -1,0 +1,49 @@
+#TABLE 2: Adjusted and unadjusted logistic regression
+# Data Preparation
+tablereg <- ofi %>% 
+  select(Sex, Age, Intubation, RTS, ISS,TimeFCT, OnDuty, daysinICU, TimeFCT, 
+         ASApreinjury, Survival, OpportunityForImprovement1)
+
+tablereg$Intubation <- ifelse(is.na(tablereg$Intubation), "Unknown", table1$Intubation)
+tablereg$Intubation <- fct_relevel(tablereg$Intubation, "No intubation", "Intubation 1-7 days", "Intubation > 7 days", "Unknown")
+
+
+# Unadjusted Table
+table3a <- tbl_uvregression(data = tablereg,
+                            method = glm,
+                            y = OpportunityForImprovement1,
+                            method.args = list(family = binomial),
+                            label = list(
+                              RTS = "Revised Trauma Score",
+                              daysinICU = "Days in the ICU",
+                              TimeFInt = "Time to first intervention",
+                              ASApreinjury = "ASA preinjury",
+                              OnDuty = "On duty",
+                              TimeFCT = "Time to first CT"
+                            )) %>%
+  bold_labels() %>%
+  bold_p(t = 0.05) 
+
+#print(table3a)
+
+# Adjusted Table
+#Creating linear regression 
+adjusted_table <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury + Survival, family = binomial, data = tablereg) 
+
+table3b <- tbl_regression(adjusted_table,
+                          label = list(RTS = "Revised Trauma Score",
+                                       daysinICU = "Days in the ICU",
+                                       ASApreinjury = "ASA preinjury",
+                                       TimeFCT = "Time to first CT")) %>%
+  bold_labels() %>%
+  bold_p(t = 0.05)
+
+# print(table3b)
+
+# Merging Tables
+table3_merge <- tbl_merge(tbls = list(table3a, table3b),
+                          tab_spanner = c("**Unadjusted**", "**Adjusted**")) %>%
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 2. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement</div>")
+
+print(table3_merge)
+

### 2024-03-18 manuscript.Rmd
18/3

@@
 title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
-output: html_document
+output: "html_document"
 author: "Elin Sun Cao"
@@
 factors and OFI among this group of patients: gender, age, if the
-patient was intubated or not and how many days, respiratory rate,
-systolic blood pressure, Glasgow Coma Scale (GCS), Injury Severity Score
-(ISS), on call, time from arrival at the hospital until first computed
-tomography (CT), ASA preinjury and survival after 30 days.
+patient was intubated or not and how many days, Revised Trauma Score
+(RTS), Injury Severity Score (ISS), on call, time from arrival at the
+hospital until first computed tomography (CT), days in the ICU, ASA
+preinjury and survival after 30 days.
 
@@
 
+OFI was identified in 143 of 1449 patients. Revised Trauma Score, ≤ 7
+days in the ICU and death within 30 days after hospitalization were
+significantly associated with OFI in the unadjusted analysis, and
+Revised Trauma Score and ≤ 7 days in the ICU were significantly
+associated with OFI in the adjusted analysis.
+
 ## Conclusion
+
+The patient and process factors associated with OFI indicated that
+higher RTS was associated with higher odds of OFI, and ≤ 7 days in the
+ICU was associated with lower odds of OFI.
 
@@
 
+RR = respiratory rate
+
 RTS = Revised Trauma Score
+
+SBP = systolic blood pressure
 
@@
 blood pressure. It has been demonstrated to show accuracy at predicting
-death.
+death. Each of the mentioned variables get a corresponding code value
+between 0 and 4 based on the value of GCS, RR and SBP. The formula for
+RTS is RTS = (0.9368 x GCS code value) + (0.7326 x SBP code
+value) + (0.2908 x RR code value) [@20].
 
@@
 from the Karolinska University Hospital in Solna and from the Swedish
-Trauma registry. We used R for all statistical analyses [@19]. For
-patients where hospital values for systolic blood pressure, respiratory
-rate or GCS score were missing, prehospital values were used instead. A
-complete case analysis was conducted after handling missing values. We
-present sample characteristics using descriptive statistics.
+Trauma registry. We used R for all statistical analyses [@19].
+
+For the RTS-value the GCS, SBP and RR values were converted to a
+RTS-code between 0 and 4.
+
+For patients where hospital values for systolic blood pressure,
+respiratory rate or GCS score were missing, prehospital values were used
+instead. If the patient was intubated A complete case analysis was
+conducted after handling missing values. We present sample
+characteristics using descriptive statistics.
 
@@
 review and specific audit filters by specialized nurses. The patients
-who are identified as having higher probability for OFI are discussed in a multidisciplinary conference held every 6-8 weeks. Examples of OFI
+who are identified as having higher probability for OFI are discussed in
+a multidisciplinary conference held every 6-8 weeks. Examples of OFI
 include the need for better organization or more senior members taking
-part in the trauma care. The presence or absence of OFI is decided among the participants of the conference and then recorded in the trauma care quality database.
+part in the trauma care. The presence or absence of OFI is decided among
+the participants of the conference and then recorded in the trauma care
+quality database.
 
@@
 categorical factors were sex, age, if the patient was intubated or not
-and how many days, respiratory rate, systolic blood pressure, Glasgow
-Coma Scale (GCS), Injury Severity Score (ISS), on call, time from
-arrival at the hospital until first computed tomography (CT), ASA
-preinjury and survival after 30 days. On duty was defined as before 8.00
-a.m., after 5 p.m., or during a weekend - defined as Saturday or Sunday.
+and how many days, Revised Trauma Score (RTS), Injury Severity Score
+(ISS), on call, time from arrival at the hospital until first computed
+tomography (CT), days in the ICU, ASA preinjury and survival after 30
+days. On duty was defined as before 8.00 a.m., after 5 p.m., or during a
+weekend - defined as Saturday or Sunday.
 
@@
 There were 14,022 patients in the trauma registry between January 1 2013
-and February 28 2023. 2679 of these patients were 
-admitted to the ICU. 3 patients were then excluded as they were younger
-than 15 years and 6 patients were excluded because they were dead on
-arrival. 928 patients were excluded because there was no registration on whether there was an OFI or not. 293 patients were excluded because they had missing values in one or more variables.
-This leaves us with 143 outcomes for OFI, and 1,306 outcomes for "No OFI".
-
-#Detta blir upprepning då det redan står längre ner i resultatdelen?
+and February 28 2023. 2679 of these patients were admitted to the ICU. 3
+patients were then excluded as they were younger than 15 years and 6
+patients were excluded because they were dead on arrival. 928 patients
+were excluded because there was no registration on whether there was an
+OFI or not. 293 patients were excluded because they had missing values
+in one or more variables. This leaves us with 143 outcomes for OFI, and
+1,306 outcomes for "No OFI".
 
@@
 As described previously, all patients with trauma team activation or
-high ISS scores are included in the database used for this study.
-Having suffered severe trauma might not be something the patients or their relatives would want to be public information as the collected data contain for instance vital parameters and demise as well as type of trauma and time of arrival to the hospital, which might be sensitive
+high ISS scores are included in the database used for this study. Having
+suffered severe trauma might not be something the patients or their
+relatives would want to be public information as the collected data
+contain for instance vital parameters and demise as well as type of
+trauma and time of arrival to the hospital, which might be sensitive
 information. Therefore, the ethical aspect of voluntary participation
@@
 # Results
+
 ## Participants
+
 ```{r, include=FALSE}
@@
 ```
-```{r}
+
+```{r, echo=FALSE}
 for (i in 1:(length(vert) - 1)) {
@@
 ## Descriptive data
-```{r include_scripts, echo=FALSE, include=FALSE}
+
+```{r include_scripts, include=FALSE}
 source("functions/table1.R")
 ```
-```{r}
+
+```{r, echo=FALSE}
 table2
@@
 ## Main results
+
 ```{r, include=FALSE}
@@
 ```
-```{r}
+
+```{r, echo=FALSE}
 table3_merge
@@
 because they were dead on arrival, and 928 patients were excluded
-because there was no data on the presence or absence OFI.
+because there was no data on the presence or absence OFI. 293 patients
+were excluded because there was missing data, leaving a total of 1,449
+patients included in the study.
 
-293 patients were excluded because there was missing data, leaving a
-total of 1,449 patients included in the study.
-
-Table 1 present sample characteristics.
-1,096 (76%) patients were male and 353 (24%) patients were female.
-846 (58%) patients had no intubation, 241 (17%) patients were intubated for 1-7 days and 113 (7,8%) patients were intubated for more than 7 days.
-The number of patients who arrived outside working hours — on Saturday or Sunday or after 5 pm or before 8 am were 1,042 (72%).
-1,064 (73%) patients were equal to or less than 7 days in the ICU, and 385 (27%) patients were more than 7 days in the ICU.
-1,080 (75%) patients had an ASA-score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA-score of 3-7 before arrival to the hospital.
-286 (20%) patients died within 30 days after hospitalization.
+Table 1 present sample characteristics. 1,096 (76%) patients were male
+and 353 (24%) patients were female. 846 (58%) patients had no
+intubation, 241 (17%) patients were intubated for 1-7 days and 113
+(7,8%) patients were intubated for more than 7 days. The number of
+patients who arrived outside working hours --- on Saturday or Sunday or
+after 5 pm or before 8 am were 1,042 (72%). 1,064 (73%) patients were
+equal to or less than 7 days in the ICU, and 385 (27%) patients were
+more than 7 days in the ICU. 1,080 (75%) patients had an ASA-score of
+1-2 before arrival to the hospital, and 369 (25%) patients had an
+ASA-score of 3-7 before arrival to the hospital. 286 (20%) patients died
+within 30 days after hospitalization.
 
@@
 analysis the factors that were significantly associated with the
-presence of OFI were: Revised Trauma Score `r inline_text(table3a, variable = RTS)`,
-≤ 7 days in the ICU `r inline_text(table3a, variable = daysinICU, level = "≤ 7 days")` and death within 30 days after hospitalization `r inline_text(table3a, variable = Survival, level = "Dead")`.
+presence of OFI were: Revised Trauma Score
+`r inline_text(table3a, variable = RTS)`, ≤ 7 days in the ICU
+`r inline_text(table3a, variable = daysinICU, level = "≤ 7 days")` and
+death within 30 days after hospitalization
+`r inline_text(table3a, variable = Survival, level = "Dead")`.
 
 In the adjusted analysis the following factors demonstrated significant
-association with OFI: Revised Trauma Score `r inline_text(table3b, variable = RTS)` and ≤ 7 days in the ICU `r inline_text(table3b, variable = daysinICU, level = "≤ 7 days")`.
+association with OFI: Revised Trauma Score
+`r inline_text(table3b, variable = RTS)` and ≤ 7 days in the ICU
+`r inline_text(table3b, variable = daysinICU, level = "≤ 7 days")`.
 
-RTS and ≤ 7 days in the ICU were significantly associated with increased odds of OFI in both the
-unadjusted and adjusted analyses.
+RTS was significantly associated with increased odds of OFI in both the
+unadjusted and adjusted analyses. ≤ 7 days in the ICU was significantly
+associated with decreasing odds of OFI in both the unadjusted and
+adjusted analyses.
 
-Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury were not significantly associated with OFI in the unadjusted or the adjusted analyses.
-Female sex, intubation > 7 days, > 7 days in the ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could not be analysed with logistic regression.
+Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury
+were not significantly associated with OFI in the unadjusted or the
+adjusted analyses. Female sex, intubation \> 7 days, \> 7 days in the
+ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could
+not be analysed with logistic regression.
 
@@
 
-We found that RTS and ≤ 7 days in the ICU were associated with higher odds of OFI. Previous
-research has shown that ...
+We found that higher RTS was associated with higher odds of OFI. RTS is
+a combined value of systolic blood pressure, Glasgow Coma Scale and
+respiratory rate with more weight towards GCS as mentioned in the
+introduction. An explanation for the increased odds of OFI in patients
+with increased RTS could be that the patients with low RTS values - with
+low GCS, RR and SBP also are the once that are in a bad state upon
+arrival - upon which emergency procedures e.g. intubation have been
+carried out rapidly due to the patients being very ill from the
+beginning. The patients who are being missed and where certain
+procedures take longer time before being performed are the patients who
+seem stable upon arrival, but who later become worse --- the patients
+with higher RTS scores. In Albaaj's study respiratory rate \>29,
+systolic blood pressure 50-75 mmHg and 76-89 mmHg, and Glasgow Coma
+Scale 9-12 were significantly related to increased odds of OFI in the
+unadjusted analysis, but not in the adjusted analysis [@12]. All of
+these variables equal medium RTS values.
 
-These findings could be because ...
+The variable ≤ 7 days in the ICU was significantly associated with
+decreasing odds of OFI in both the adjusted and unadjusted analyses. An
+explanation would be that less days spent in the ICU compared to more
+days spent in the ICU indicates that patients have been in a better
+state and needed less time of recovery before leaving the hospital.
+Furthermore, a longer stay leads to more space for mistakes being made.
 
-Death within 30 days after hospitalisation was associated with OFI in the unadjusted analysis but not in the
-adjusted analysis.
-
-1.  add sources from previous studies
-
-2.  studies to back up the results from the analyses
+Death within 30 days after hospitalization was associated with OFI in
+the unadjusted analysis but not in the adjusted analysis. Patients who
+died within 30 days after hospitalization could either have suffered a
+severe trauma or have been in a poor state from the beginning; therefore
+increasing the odds of OFI.
 
@@
 
-There were several limitations in our study. The selection for review
-for the morbidity and mortality review relied mostly on the local audit
-filters used at the Karolinska Institute Hospital in Solna. This leads
-to a risk of misplacement where some patients with OFI were not selected
-by for the review based on the audit filters although there was an OFI.
+There were several limitations in our study: the selection for the
+morbidity and mortality review, the definition of the RTS variable and
+the study size.
 
-Another limtation is that some relevant factors, such as ----, were not
-part of the Swedish Trauma registry and therefore not adjusted for in
-the analysis.
+The selection for review for the morbidity and mortality review relied
+mostly on the local audit filters used at the Karolinska Institute
+Hospital in Solna. This leads to a risk of misplacement where some
+patients with OFI were not selected by for the review based on the audit
+filters although there was an OFI later on in the hospitalization
+process. #Lägga in källa för audit filters - Kelvins artikel?
 
-The study was a single-center study and illustrate the situation in
-Stockholm. More study is needed on the area of OFI in patients admitted
-to the ICU. A suggestion would be to conduct a multi-center study on
-this group of patients to further investigate areas for improvement and
-to further develop trauma quality programs.
+The definition of the RTS-value is a combination of GCS, SBP and RR.
+When the patients were intubated and had missing values for GCS or RR,
+we replaced it with RTS with a value of 0 for the equation for RTS, as
+GCS and respiratory rate are 0 when patients are intubated. This could
+affect the final results, because then we would not get their initial
+values for these variables, meaning some patients received a RTS score
+more weighted towards SBP than they should have. However, the use of RTS
+can also be seen as an advantage, being a combination of vital
+parameters. This would mean that the RTS value reflects the state of the
+patient better than these parameters separately.
 
-Future studies could also focus on ---- which was seen to be associated
-with OFI in these patients.
-
-<!--Lägg till 2 st förslag på explicita fortsättningsstudier! -->
+The study was a single-centre study and illustrate the situation in the
+Karolinska Intitute Hospital in Solna. More research is needed on the
+area of OFI in patients admitted to the ICU. A suggestion would be to
+conduct a multi-centre study on this group of patients to further
+investigate areas for improvement and to further develop trauma quality
+programs. Based on the results from this study, variables of interest
+could be number of days in the ICU and investigation of GCS, RR and SBP
+separately instead of them as a combined RTS-score. A study including
+more patients could potentially show significance for the separate vital
+parameters leading to a better prediction model for which patients that
+should receive more focus when improving trauma care.
 
@@
 
-The study is conducted on 143 outcomes of OFI. There were few outcomes because
-we included only those who were admitted to the ICU, \> 14 years old,
-not dead on arrival and had a registration for the presence or absence
-of OFI. The external validity is therefore compromised due to the
-limited size of the Swedish database. However, ...
+The study is conducted on 143 outcomes of OFI. There were few outcomes
+because we included only those who were admitted to the ICU, \> 14 years
+old, not dead on arrival and had a registration for the presence or
+absence of OFI. The external validity is therefore compromised due to
+the limited size of the Swedish database.
+
+Based on the statistical significance for higher RTS being associated
+with higher odds of OFI, there should be more focus on trauma patients
+who are stable upon arrival who have a risk of becoming worse. Future
+studies should investigate what factors to use to see the indication for
+such a focus. Focusing on patients who are stable upon arrival overall
+is very extensive. Therefore, identifying such factors would be
+valuable.
 
@@
 
----- among patients admitted to the ICU have higher odds of OFI. These
-factors can be seen as new targets for further reseach to further
-supplement trauma quality improvement programs.
-
-<!-- Give a short conclusion, corresponding to the aim -->
+The aim of this study was to characterize patient and process factors
+among adult patients admitted to the ICU that were associated with OFI.
+The study showed that higher RTS score is associated with increased odds
+of OFI indicating that patients who are relatively well upon arrival to
+the hospital are the group where improved trauma care should be focused
+on. The logistical regression also showed that ≤ 7 days in the ICU was
+associated with decreased odds of OFI indicating that patients who are
+at the hospital less days in the ICU compared to more days in the ICU
+are less prone to opportunities for improvement.
 

### 2024-03-18 functions/table1.R

@@ -16,6 +16,7 @@ library(dplyr)
 #install.packages("gtsummary")
 library(gtsummary)
 library(tidyverse)
+library(officer)
 
 
 #FLOWCHART: included/excluded
@@ -115,15 +116,15 @@ ofi$RTSSBP <- ifelse(ofi$SystolicBloodPressure > 89, 4,
                                           ifelse(ofi$SystolicBloodPressure == 0, 0,
                                                  ifelse(ofi$SystolicBloodPressure == 99, 0, NA))))))
 
-ofi$RTSRR <- ifelse(ofi$RespiratoryRate > 29, 4,
-                    ifelse(ofi$RespiratoryRate %in% 10:29, 3,
+ofi$RTSRR <- ifelse(ofi$RespiratoryRate %in% 10:29, 4,
+                    ifelse(ofi$RespiratoryRate >29, 3,
                            ifelse(ofi$RespiratoryRate %in% 6:9, 2,
                                   ifelse(ofi$RespiratoryRate %in% 1:5, 1,
                                          ifelse(ofi$RespiratoryRate == 0, 0,
                                                 ifelse(ofi$RespiratoryRate == 99, 0, NA)))))) 
 
 ofi$RTS <- (0.9368*ofi$RTSGCS + 0.7326*ofi$RTSSBP + 0.2908*ofi$RTSRR)
-
+#ofi$RTS <- (ofi$RTSGCS + ofi$RTSSBP + ofi$RTSRR)
 
 
 
@@ -172,6 +173,8 @@ ofi$OpportunityForImprovement1 <- ifelse(ofi$OpportunityForImprovement == "Oppor
 #Creating new table with defined data 
 library(dplyr)
 library(gt)
+#install.packages("flextable")
+library(flextable)
 
 table1 <- ofi %>% 
   select(Sex, Age, Intubation, RTS, ISS, TimeFCT, OnDuty, daysinICU, 
@@ -207,4 +210,5 @@ table2 <- table1 %>%
   bold_labels() %>% 
   add_overall(last = TRUE) %>% 
   modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 1. Sample Characteristics</div>") %>% 
+ # as_flex_table() %>%
   print()

### 2024-03-18 functions/table2.R

@@ -13,6 +13,7 @@ table3a <- tbl_uvregression(data = tablereg,
                             method = glm,
                             y = OpportunityForImprovement1,
                             method.args = list(family = binomial),
+                            exponentiate = TRUE, 
                             label = list(
                               RTS = "Revised Trauma Score",
                               daysinICU = "Days in the ICU",
@@ -31,6 +32,7 @@ table3a <- tbl_uvregression(data = tablereg,
 adjusted_table <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury + Survival, family = binomial, data = tablereg) 
 
 table3b <- tbl_regression(adjusted_table,
+                          exponentiate = TRUE, 
                           label = list(RTS = "Revised Trauma Score",
                                        daysinICU = "Days in the ICU",
                                        ASApreinjury = "ASA preinjury",

### 2024-03-19 manuscript.Rmd
19/3

@@
 
-This was a registry-based study using all patients between 2014 and 2022
+This was a registry-based study using all patients between 2014 and 2023
 from the Karolinska University Hospital in Solna who had been admitted
@@
 
-OFI was identified in 143 of 1449 patients. Revised Trauma Score, ≤ 7
-days in the ICU and death within 30 days after hospitalization were
-significantly associated with OFI in the unadjusted analysis, and
-Revised Trauma Score and ≤ 7 days in the ICU were significantly
-associated with OFI in the adjusted analysis.
+OFI was identified in 143 of 1449 patients. Higher Revised Trauma Score
+(RTS) was associated with higher odds of OFI, ≤ 7 days in the intensive
+care unit (ICU) was associated with lower odds of OFI, and death within
+30 days after hospitalization was associated with higher odds of OFI
+with statistical significance in the unadjusted analysis. Higher Revised
+Trauma Score (RTS) was associated with higher odds of OFI, and ≤ 7 days
+in the ICU was associated with lower odds of OFI with statistical
+significance in the adjusted analysis.
 
@@
 
-For the RTS-value the GCS, SBP and RR values were converted to a
-RTS-code between 0 and 4.
-
 For patients where hospital values for systolic blood pressure,
 respiratory rate or GCS score were missing, prehospital values were used
-instead. If the patient was intubated A complete case analysis was
-conducted after handling missing values. We present sample
-characteristics using descriptive statistics.
+instead. If the patient was intubated we automatically assumed that GCS
+was equal to 3 and RR was equal to 0. For the RTS-variable, each of the
+variables GCS, SBP and RR were then converted to a RTS-code between 0
+and 4 based on what the values of GCS, SBP and RR were respectively. RTS
+was then calculated using the formula for RTS based on the combination
+of the RTS-codes assigned to each of the previously listed variables.
+
+A complete case analysis was conducted after handling missing values. We
+present sample characteristics using descriptive statistics.
 
@@
 a combined value of systolic blood pressure, Glasgow Coma Scale and
-respiratory rate with more weight towards GCS as mentioned in the
-introduction. An explanation for the increased odds of OFI in patients
-with increased RTS could be that the patients with low RTS values - with
-low GCS, RR and SBP also are the once that are in a bad state upon
-arrival - upon which emergency procedures e.g. intubation have been
-carried out rapidly due to the patients being very ill from the
-beginning. The patients who are being missed and where certain
-procedures take longer time before being performed are the patients who
-seem stable upon arrival, but who later become worse --- the patients
-with higher RTS scores. In Albaaj's study respiratory rate \>29,
-systolic blood pressure 50-75 mmHg and 76-89 mmHg, and Glasgow Coma
-Scale 9-12 were significantly related to increased odds of OFI in the
-unadjusted analysis, but not in the adjusted analysis [@12]. All of
-these variables equal medium RTS values.
+respiratory rate with more weight towards GCS. An explanation for the
+increased odds of OFI in patients with increased RTS could be that the
+patients with low RTS values - with low GCS, RR and SBP also are the
+once that are in a bad state upon arrival --- upon which emergency
+procedures e.g. intubation have been carried out rapidly due to the
+patients being visibly ill from the beginning. The patients who are
+being missed and where certain procedures take longer time before being
+performed are the patients who seem stable upon arrival, but who later
+become worse --- the patients with higher RTS scores. In Albaaj's study
+respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
+mmHg, and Glasgow Coma Scale 9-12 were significantly related to
+increased odds of OFI in the unadjusted analysis, but not in the
+adjusted analysis among patients from the Swedish Trauma Registry [@12].
+All of these variables equal medium to higher RTS values --- capturing
+the group of patients who's vital parameters are negatively affected by
+the trauma, but who are relatively stable in this group of patients.
+This creates a sense of false security, leading to increased time until
+necessary emergency procedures are performed.
+
+#Se om det finns källa som pekar på vilken grupp av patienter som bör få
+mer fokus på IVA?
 
 The variable ≤ 7 days in the ICU was significantly associated with
-decreasing odds of OFI in both the adjusted and unadjusted analyses. An
-explanation would be that less days spent in the ICU compared to more
-days spent in the ICU indicates that patients have been in a better
-state and needed less time of recovery before leaving the hospital.
-Furthermore, a longer stay leads to more space for mistakes being made.
+decreasing odds of OFI in both the adjusted and unadjusted analyses.
+There are two aspects to consider concerning this variable. One aspect
+would be that less days spent in the ICU compared to more days spent in
+the ICU indicates that patients have been in a better state and needed
+less time of recovery before leaving the hospital. Furthermore, a longer
+stay leads to more space for mistakes being made. The other aspect of
+the variable would be that few days in the ICU could indicate the
+patient passing away before any mistakes having been made due to the
+patient being very ill from the beginning --- the deaths were not
+preventable, but rather expected.
+
+#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
+vad orsakerna är
 
@@
 increasing the odds of OFI.
+
+Sex, age and ASA preinjury were not significant in the unadjusted nor
+the adjusted analyses, which implicates that the patients were not
+treated differently based on age or gender. The results concerning the
+ASA preinjury could be due to the OFI considering if each patient really
+had an opportunity for improvement based on their individual cases,
+meaning that if a patient has a higher ASA-score before the trauma, that
+patient may not have the same reference point for OFI compared to a
+patient with low ASA-score before the trauma.
+
+#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
+ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
+bedömningen av OFI
+
+The variable on duty was not significant in either analyses indicating
+that the time of arrival of the patients did not affect their medical
+treatment. Trying to reduce number of OFI can therefore not be solved by
+increasing medical resources outside working hours.
+
+#Lägga in artikel som pekar på om duty påverkar handläggning eller inte
+
+Something to consider is why ISS was not significant in the analyses.
+Patients admitted to the ICU are usually in a bad shape with generally
+high ISS. ISS is based on the square of the three highest AIS scores
+allocated to the three most severly injured body parts. The downside of
+ISS on its own is that it has higher predictive value when combined with
+other parameters like vital signs, comorbidities and mechanism of
+injury.
+
+#Artikel om hur ISS bedöms kunna användas i samanhang av trauma
 
@@
 filters although there was an OFI later on in the hospitalization
-process. #Lägga in källa för audit filters - Kelvins artikel?
+process.
+
+#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)
 
@@
 absence of OFI. The external validity is therefore compromised due to
-the limited size of the Swedish database.
+the limited size of the Swedish trauma database of patients admitted to
+the ICU.
 
@@
 of OFI indicating that patients who are relatively well upon arrival to
-the hospital are the group where improved trauma care should be focused
-on. The logistical regression also showed that ≤ 7 days in the ICU was
-associated with decreased odds of OFI indicating that patients who are
-at the hospital less days in the ICU compared to more days in the ICU
-are less prone to opportunities for improvement.
+the hospital are the group where improved trauma care should get more
+focus than they do now. The logistical regression also showed that ≤ 7
+days in the ICU was associated with decreased odds of OFI indicating
+that patients who are at the hospital less days in the ICU compared to
+more days in the ICU are less prone to opportunities for improvement.
 

### 2024-03-20 manuscript.Rmd
20/3

@@
 performed are the patients who seem stable upon arrival, but who later
-become worse --- the patients with higher RTS scores. In Albaaj's study
-respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
-mmHg, and Glasgow Coma Scale 9-12 were significantly related to
-increased odds of OFI in the unadjusted analysis, but not in the
+become worse --- the patients with comparatively higher RTS scores. In
+Albaaj's study respiratory rate \>29, systolic blood pressure 50-75 mmHg
+and 76-89 mmHg, and Glasgow Coma Scale 9-12 were significantly related
+to increased odds of OFI in the unadjusted analysis, but not in the
 adjusted analysis among patients from the Swedish Trauma Registry [@12].
@@
 the group of patients who's vital parameters are negatively affected by
-the trauma, but who are relatively stable in this group of patients.
-This creates a sense of false security, leading to increased time until
-necessary emergency procedures are performed.
+the trauma, but who are relatively stable in the group of trauma
+patients. This creates a sense of false security, leading to increased
+time until necessary emergency procedures are performed.
 
-#Se om det finns källa som pekar på vilken grupp av patienter som bör få
-mer fokus på IVA?
+A previous study on pediatric patients in South Africa showed that RTS
+was not predictive of mortality between high-risk and low-risk patients
+[@21]. This could indicate... .
 
@@
 
-#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
-vad orsakerna är
+Previous studies have shown that prolonged ICU stay is a risk factor for
+increased mortality and morbidity in trauma patients [@22, @26].
+Something to consider is the reason for the lengthened stay in the ICU -
+studies have shown that patient characteristics on admission, like
+co-morbidities, age and severity of illness do not fully explain the
+increased long-term mortality. With lengthened stay in the ICU the
+reason for stay in the ICU become increasingly less associated with
+patient's original reason for admittance and increasingly associated
+with persistent chronic illness with associated organ dysfunction [@23,
+@24].
+
+This being the case, this might be an explanation for the lack of
+sigificance for ASA preinjury and ISS in the analysis. As stated above
+severity of illness do not fully explain increased long-term mortality.
+OFI may therefore not correlate fully with ASA preinjury and ISS
+respectively.
+
+However, a retrospective cohort has also shown that lower admission
+illness severity and younger age were associated with earlier discharge
+and therefore less associated with persistent critical illness, which
+would then indicate a stronger association between ISS and ASA preinjury
+to length of stay in the ICU as the patients who are in a better state
+leave the hospital earlier and there is less possibilities for OFI. On
+the other hand, the cohort also showed that lower comorbidity was
+associated with persistent critical illness rather than early discharge
+from the ICU --- supporting the opposite side of the argument [@25].
+
+Several studies have also studied the association between different
+factors which affect the length of stay in the ICU. Associations between
+hypoglycaemia, use of any corticosteroids, use of any neuromuscular
+agents, use of any benzodiazepines for more than 1 day, the need for
+mechanical ventilation for \> 2 days, initiation of new dialysis and the
+occurrence of new infection has been found with increased total and
+5-year mortality [@23]. An association between bleeding in patients and
+increased length of stay in the ICU [@27].
+
+KLADD: APACHE is a physiological score and works best on critically ill
+patients on a group level, but not as well on individual patients or on
+relatively healthier patients [@28]...
 
@@
 
-#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
-ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
-bedömningen av OFI
+There are many studies on gender in association to health care. It has
+previously been shown that there is no difference in willingness to
+admit to the ICU between cases describing a man or woman in the
+physician responses [@29]. However, studies also show that, among
+severely injured patients, female gender is associated with lower
+in-hospital mortality among those aged 16-44 years. Female gender is
+also associated with decreased likelihood for ICU admission [@30]. Male
+gender contributes more to poor outcomes in postoperative ICU patients,
+and two-thirds of ICU resources are consumed by male patients [@31].
+However, in our study we could not see any correlation between OFI and
+one specific gender.
+
+Preinjury ASA-PS score has been shown to be a reliable score for
+predicting mortality after trauma, predicting readmission after
+traumatic injury and for classifying comorbidity in trauma patients
+[@32, @33, @34]. But why is ASA preinjury not associated with OFI? Could
+it be because longer ICU stays are not associated with comorbidities to
+the same extent? ...
 
@@
 treatment. Trying to reduce number of OFI can therefore not be solved by
-increasing medical resources outside working hours.
-
-#Lägga in artikel som pekar på om duty påverkar handläggning eller inte
+increasing medical resources outside working hours. On-call sleeping has
+been studied and it has been stated that sleeping between calls is
+beneficial for waking performance. However, there might be a temporary
+increase in performance impairment that should be taken into
+consideration [@35]...
 
@@
 
-#Artikel om hur ISS bedöms kunna användas i samanhang av trauma
+ISS is an anatomical model, compared to RTS which is a physiological
+model. Among trauma patients admitted to the ICU the physiological
+models perform better than the anatomical models. However, patients with
+low scores in the physiological model also require an anatomic analysis
+of the injuries to determine their severity [@36]. NISS has been found
+to be better than ISS in predicting the ICU admission of trauma
+patients, as well as predicting prolonged ICU length of stay [@37]. This
+could be an explanation to why ISS is not significant in the results of
+this analysis, as RTS is a better model for patients admitted to the
+ICU.
 
-### Limitations
+### Strengths and limitations
+
+The strengths of this study was that it is a national registry-based
+study which means that the quality of the data is reliable...
 
@@
 mostly on the local audit filters used at the Karolinska Institute
-Hospital in Solna. This leads to a risk of misplacement where some
+Hospital in Solna as a base for choosing patient cases for the mortality
+and morbidity review. Using audit filters to select patients for the
+conference is associated with high false positive rates, from 24% to 80%
+[@38]. Even though the use of audit filters is associated with high
+false positive rates, there is still a risk of misplacement where some
 patients with OFI were not selected by for the review based on the audit
-filters although there was an OFI later on in the hospitalization
-process.
-
-#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)
+filters although there was an OFI that occurred later on during the
+hospital stay.
 
@@
 
+Another limitation is the combination of data from the emergency
+department and prehospital values when the previous one's were missing.
+
 The study was a single-centre study and illustrate the situation in the
@@
 should receive more focus when improving trauma care.
+
+Examples of limitations:
+
+-   Theoretical: limits the scope, depth, or applicability of a study.
+
+-   Methodological: limits the quality, quantity, or diversity of the
+    data.
+
+-   Empirical: limits the representativeness, validity, or reliability
+    of the data.
+
+-   Analytical: limits the accuracy, completeness, or significance of
+    the findings.
+
+-   Ethical: limits the access, consent, or confidentiality of the data.
 

### 2024-03-25 manuscript.Rmd
25/3

--- 2024-03-19 manuscript.Rmd
@@
 # Abstract
+
+## Bakgrund
+
+Trauma är en ledande orsak till mortalitet och morbiditet i världen. En
+hörnsten i utvecklingen av förbättringsprogram för trauma är
+multidisciplinära mortalitets- och morbiditetskonferenser. Syftet med
+dessa är att identifiera opportunities for improvement (OFI) som är en
+viktig del i det riktade förbättringsarbetet av traumahandläggning.
+Många patienter med svåra trauman blir inlagda på
+intensivvårdsavdelningen (IVA). Trots detta är kännedomen om
+förbättringsmöjligheter hos denna patientgrupp inte heltäckande i
+nuläget. Syftet med denna studie är att karaktärisera OFI hos vuxna
+traumapatienter som varit inlagda på IVA och bedöma hur patient- och
+processfaktorer är associerade med OFI hos denna patientgrupp.
+
+## Metod
+
+Detta är en registerbaserad studie som inkluderar traumapatienter från
+2014 till 2023 från Karolinska Universitetssjukhuset i Solna som har
+varit inlagda på IVA och som har granskats avseende förekomsten av OFI
+vid en mortalitets- och morbiditetskonferens. Bivariabel och
+multivariabel logistisk regression användes för att bedöma sambandet
+mellan följande patient- och processfaktorer samt OFI bland denna
+patientgrupp: kön, ålder, om patienten var intuberad eller ej och hur
+många dagar, Revised Trauma Score (RTS), Injury Severity Score (ISS),
+jourtid vid ankomst till sjukhuset, tid från ankomst till sjukhuset till
+första datortomografi (DT), dagar på IVA, ASA-värde före trauma och
+överlevnad efter 30 dagar.
+
+## Resultat
+
+OFI identifierades hos 143 av 1449 patienter. Högre Revised Trauma Score
+(RTS) var associerat med högre odds för OFI, ≤ 7 dagar på
+intensivvårdsavdelningen (IVA) var associerat med lägre odds för OFI och
+död inom 30 dagar efter sjukhusvistelse var associerat med högre odds
+för OFI med statistisk signifikans i den ojusterade analysen. Högre
+Revised Trauma Score (RTS) var associerat med högre odds för OFI, och ≤
+7 dagar på IVA var associerat med lägre odds för OFI med statistisk
+signifikans i den justerade analysen.
+
+## Slutsats
+
+De patient- och processfaktorer som var signifikant associerade med OFI
+visade att högre RTS var associerat med högre odds för OFI och att ≤ 7
+dagar på IVA var associerat med lägre odds för OFI.
 
@@
 to the ICU and who were reviewed regarding the presence of OFI by the
-mortality and morbidity conference. Bi-variable and multivariable
+mortality and morbidity conference. Bivariable and multivariable
 logistic regression was used to assess the following patient and process
@@
 patient was intubated or not and how many days, Revised Trauma Score
-(RTS), Injury Severity Score (ISS), on call, time from arrival at the
-hospital until first computed tomography (CT), days in the ICU, ASA
-preinjury and survival after 30 days.
+(RTS), Injury Severity Score (ISS), arrival outside of working hours,
+time from arrival at the hospital until first computed tomography (CT),
+days in the ICU, ASA preinjury and survival after 30 days.
 
@@
 # Abbreviations
+
+APACHE = Acute Physiology and Chronic Health Evaluation
+
+ASA score = American Society of Anesthesiologists score
 
@@
 Albaaj et al. recently analyzed opportunities for improvement (OFI) in
-relation to different patient and process factors based on data from the
-Swedish trauma registry and data from the Karolinska University Hospital
-in Solna. The study indicated that patients with moderate to severe
-trauma are at the highest odds of OFI [@4]. According to SweTrau 2022
-41,1% of patients with New Injury Severity Score (NISS) \> 15 who were
-at a university hospital were admitted to the ICU. The proportion of
-severely injured admitted to the ICU has decreased in recent years,
-however; patients with NISS \> 15 still constitute a large proportion of
-the trauma patients treated in the ICU. A way to measure functional
-level after hospital discharge is Glasgow Outcome Scale (GOS). According
-to GOS, among patients with NISS \> 15 17% were deceased, 20% suffered
-severe disability, 49% suffered moderate disability and 14% had good
-recovery in the university hospitals of Sweden [@3].
+relation to different patient and process factors based on a selection
+of data from SweTrau including the patients admitted to the Karolinska
+University Hospital in Solna following trauma. The study indicated that
+patients with moderate to severe trauma are at the highest odds of OFI
+[@4]. According to SweTrau 2022 41,1% of patients with New Injury
+Severity Score (NISS) \> 15 who were at a university hospital were
+admitted to the ICU. The proportion of severely injured admitted to the
+ICU has decreased in recent years, however; patients with NISS \> 15
+still constitute a large proportion of the trauma patients treated in
+the ICU. A way to measure functional level after hospital discharge is
+Glasgow Outcome Scale (GOS). According to GOS, among patients with NISS
+\> 15 17% were deceased, 20% suffered severe disability, 49% suffered
+moderate disability and 14% had good recovery in the university
+hospitals of Sweden [@3].
 
@@
 
+## Classifications systems for patient's general health
+
+Another way to assess the state of a patient in addition to the injury
+score systems are classification systems for patient's general health.
+Two of these are ASA score and APACHE II.
+
+American Society of Anesthesiologists (ASA) classification was
+originally created to establish a scoring system for evaluation of a
+patient's general health and comorbidities before an operative procedure
+[@39]. However, recent studies have also showed ASA class to be a
+significant predictor for length of stay, indicating that ASA score can
+be used as predictor in other aspects of hospital care as well [@40].
+
+Acute Physiology and Chronic Health Evaluation (APACHE) II is the most
+commonly used severity-of-disease scoring system in ICUs all around the
+world. The worst value for each physiological variabel is converted to
+an integer score from 0 to 71 within the first 24 hours of admittance.
+Higher score indicates more severe disease and higher risk for
+mortality. The score has not been validated for use on patient's under
+16 years of age. APACHE II has shown to be a good score for
+discriminating between survivors and non-survivors [@41].
+
 ## Preventable deaths in patients admitted to the ICU
@@
 For patients where hospital values for systolic blood pressure,
-respiratory rate or GCS score were missing, prehospital values were used
-instead. If the patient was intubated we automatically assumed that GCS
-was equal to 3 and RR was equal to 0. For the RTS-variable, each of the
-variables GCS, SBP and RR were then converted to a RTS-code between 0
-and 4 based on what the values of GCS, SBP and RR were respectively. RTS
-was then calculated using the formula for RTS based on the combination
-of the RTS-codes assigned to each of the previously listed variables.
+respiratory rate or GCS score were missing from the emergency
+department, prehospital values were used instead. If the patient was
+intubated we automatically assumed that GCS was equal to 3 and RR was
+equal to 0. For the RTS-variable, each of the variables GCS, SBP and RR
+were then converted to a RTS-code between 0 and 4 based on what the
+values of GCS, SBP and RR were respectively. RTS was then calculated
+using the formula for RTS based on the combination of the RTS-codes
+assigned to each of the previously listed variables.
 
@@
 
-The study outcome will be on the presence of OFI in adult patients
-treated in the ICU based on patient and process factors.
+The study outcome is on the presence of OFI in adult patients treated in
+the ICU based on patient and process factors.
 
@@
 
-The quantitative variables were factors chosen from the trauma registry,
-based on the locally used audit filters, standard epidemiological
-factors and factors registered in the Swedish Trauma Registry. The
-categorical factors were sex, age, if the patient was intubated or not
-and how many days, Revised Trauma Score (RTS), Injury Severity Score
-(ISS), on call, time from arrival at the hospital until first computed
-tomography (CT), days in the ICU, ASA preinjury and survival after 30
-days. On duty was defined as before 8.00 a.m., after 5 p.m., or during a
-weekend - defined as Saturday or Sunday.
+The variables were factors chosen from the trauma registry, based on the
+locally used audit filters, standard epidemiological factors and factors
+registered in the Swedish Trauma registry. The categorical factors were
+sex, age, if the patient was intubated or not and how many days, Revised
+Trauma Score (RTS), Injury Severity Score (ISS), on call, time from
+arrival at the hospital until first computed tomography (CT), days in
+the ICU, ASA preinjury and survival after 30 days. On duty was defined
+as before 8.00 a.m., after 5 p.m., or during a weekend - defined as
+Saturday or Sunday.
 
@@
 There were 14,022 patients in the trauma registry between January 1 2013
-and February 28 2023. 2679 of these patients were admitted to the ICU. 3
+and February 1 2023. 2679 of these patients were admitted to the ICU. 3
 patients were then excluded as they were younger than 15 years and 6
@@
 
+14,022 patients were included in the trauma registry and trauma care
+quality database between 2017 and 2023. 12,278 patients were excluded,
+leaving a total of 1,742 patients eligible for the study. 11,343 were
+excluded because they were not treated in the ICU. 3 patients were
+excluded because they were under 15 years old. 6 patients were excluded
+because they were dead on arrival, and 928 patients were excluded
+because there was no data on the presence or absence OFI. 293 patients
+were excluded because there was missing data, leaving a total of 1,449
+patients included in the study.
+
 ## Descriptive data
@@
 ```
-
-## Main results
-
-```{r, include=FALSE}
-source("functions/table2.R")
-```
-
-```{r, echo=FALSE}
-table3_merge
-```
-
-14,022 patients were included in the trauma registry and trauma care
-quality database between 2017 and 2023. 12,278 patients were excluded,
-leaving a total of 1,742 patients eligible for the study. 11,343 were
-excluded because they were not treated in the ICU. 3 patients were
-excluded because they were under 15 years old. 6 patients were excluded
-because they were dead on arrival, and 928 patients were excluded
-because there was no data on the presence or absence OFI. 293 patients
-were excluded because there was missing data, leaving a total of 1,449
-patients included in the study.
 
@@
 within 30 days after hospitalization.
+
+## Main results
+
+```{r, include=FALSE}
+source("functions/table2.R")
+```
+
+```{r, echo=FALSE}
+table3_merge
+```
 
@@
 Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury
-were not significantly associated with OFI in the unadjusted or the
+were not significantly associated with OFI in the unadjusted nor the
 adjusted analyses. Female sex, intubation \> 7 days, \> 7 days in the
@@
 
+```{r, include=FALSE}
+source("functions/table3.R")
+```
+
+```{r, echo=FALSE}
+table3alive_merge
+```
+
+Table 3 shows the unadjusted and adjusted associations of OFI and the
+selected patient and process factors among the patients who were alive
+30 days after hospitalization. In the unadjusted analysis time to first
+CT `r inline_text(table3aalive, variable = TimeFCT)` was significant.
+Indicating that it has no correlation with increasing or decreasing odds
+of OFI.
+
 # Discussion
 
-### Key results
+The key results from this study are that we found that higher RTS was
+associated with higher odds of OFI in both the adjusted and unadjusted
+analyses, ≤ 7 days in the ICU was associated with decreased risk of OFI
+in both the unadjusted and adjusted analyses, and death within 30 days
+after hospitalization was associated with OFI in the unadjusted, but not
+in the adjusted analysis. Age, gender, intubation, ISS, time to first
+CT, on duty and ASA preinjury were not significantly associated with OFI
+in neither the unadjusted or the adjusted analyses. In the subgroup
+analysis of patients who were alive 30 days after hospitalization time
+to first CT showed significance for having no relation to OFI.
 
@@
 patients with low RTS values - with low GCS, RR and SBP also are the
-once that are in a bad state upon arrival --- upon which emergency
+ones that are in a bad state upon arrival --- upon which emergency
 procedures e.g. intubation have been carried out rapidly due to the
@@
 performed are the patients who seem stable upon arrival, but who later
-become worse --- the patients with higher RTS scores. In Albaaj's study
-respiratory rate \>29, systolic blood pressure 50-75 mmHg and 76-89
-mmHg, and Glasgow Coma Scale 9-12 were significantly related to
-increased odds of OFI in the unadjusted analysis, but not in the
-adjusted analysis among patients from the Swedish Trauma Registry [@12].
-All of these variables equal medium to higher RTS values --- capturing
-the group of patients who's vital parameters are negatively affected by
-the trauma, but who are relatively stable in this group of patients.
-This creates a sense of false security, leading to increased time until
-necessary emergency procedures are performed.
-
-#Se om det finns källa som pekar på vilken grupp av patienter som bör få
-mer fokus på IVA?
+become worse --- the patients with comparatively higher RTS scores. In
+Albaaj's study respiratory rate \>29, systolic blood pressure 50-75 mmHg
+and 76-89 mmHg, and Glasgow Coma Scale 9-12 were significantly related
+to increased odds of OFI in the unadjusted analysis, but not in the
+adjusted analysis among patients from the part of the Swedish Trauma
+Registry covering patients from the Karolinska University Hospital in
+Solna [@12]. All of these variables equal medium to higher RTS values
+--- capturing the group of patients who's vital parameters are
+negatively affected by the trauma, but who are relatively stable in the
+group of trauma patients. This creates a sense of false security,
+leading to increased time until necessary emergency procedures are
+performed.
 
@@
 patient passing away before any mistakes having been made due to the
-patient being very ill from the beginning --- the deaths were not
-preventable, but rather expected.
+patient being very ill from the beginning --- cases in which the deaths
+were not preventable, but rather expected.
 
-#Leta efter artikel om att längre vårdtis på IVA ger sämre utfall och
-vad orsakerna är
+Previous studies have shown that prolonged ICU stay is a risk factor for
+increased mortality and morbidity in trauma patients [@22, @26].
+Something to consider is the reason for the lengthened stay in the ICU.
+Studies have shown that patient characteristics on admission, like
+comorbidities, age and severity of illness do not fully explain the
+increased long-term mortality. With lengthened stay in the ICU the
+reason for stay in the ICU become increasingly less associated with
+patient's original reason for admittance and increasingly associated
+with persistent chronic illness with associated organ dysfunction [@23,
+@24]. This being the case, this might be an explanation for the lack of
+sigificance for ASA preinjury and ISS in the analysis. As stated above
+severity of illness and comorbidities do not fully explain increased
+long-term mortality. OFI may therefore not correlate fully with ASA
+preinjury and ISS respectively which are factors describing
+comorbidities and the severity of illness.
+
+However, a retrospective cohort has shown that lower admission illness
+severity and younger age were associated with earlier discharge and
+therefore less associated with persistent critical illness, which would
+then indicate a stronger association between ISS and ASA preinjury to
+length of stay in the ICU as the patients who are in a better state
+leave the hospital earlier and there is less odds for OFI. On the other
+hand, the cohort also showed that lower comorbidity was associated with
+persistent critical illness rather than early discharge from the ICU ---
+supporting the opposite side of the argument [@25].
+
+Several studies have also studied the association between different
+factors that affect the length of stay in the ICU. Associations between
+hypoglycaemia, use of any corticosteroids, use of any neuromuscular
+agents, use of any benzodiazepines for more than 1 day, the need for
+mechanical ventilation for \> 2 days, initiation of new dialysis and the
+occurrence of new infection has been found to have a correlation to
+increased total and 5-year mortality [@23]. An association between
+bleeding in patients and increased length of stay in the ICU has also
+been found [@27].
 
@@
 
-Sex, age and ASA preinjury were not significant in the unadjusted nor
-the adjusted analyses, which implicates that the patients were not
-treated differently based on age or gender. The results concerning the
-ASA preinjury could be due to the OFI considering if each patient really
-had an opportunity for improvement based on their individual cases,
-meaning that if a patient has a higher ASA-score before the trauma, that
-patient may not have the same reference point for OFI compared to a
-patient with low ASA-score before the trauma.
+Sex and age were not significant in the unadjusted nor the adjusted
+analyses, which implicates that the patients were not treated
+differently based on age or gender. There are many studies on gender in
+association to health care. It has previously been shown that there is
+no difference in willingness to admit to the ICU between cases
+describing a man or woman in the physician's responses [@29]. However,
+studies also show that, among severely injured patients, female gender
+is associated with lower in-hospital mortality among those aged 16-44
+years. Female gender is also associated with decreased likelihood for
+ICU admission [@30]. Male gender contributes more to poor outcomes in
+postoperative ICU patients, and two-thirds of ICU resources are consumed
+by male patients [@31]. However, in our study we could not see any
+correlation between OFI and one specific gender in the occurrence of
+OFI. Although we can also see that the majority of the patients included
+in this study were also males. There is possibly a reason for higher
+admittance of males to the ICU and association with poorer outcomes, but
+it has nothing to do with odds of OFI.
 
-#Lägga in artikel om hur IVA patienter behandlas baserat på kön eller
-ålder + artikel om hur ASA-poängen eller pat allvarlighetsgrad påverkar
-bedömningen av OFI
+ASA preinjury was not significant in the unadjusted nor the adjusted
+analyses. Preinjury ASA score has been shown to be a reliable score for
+predicting mortality after trauma, predicting readmission after
+traumatic injury and for classifying comorbidity in trauma patients
+[@32, @33, @34]. ASA score was originally developed to assess a
+patient's health before surgical procedures and therefore provided
+information on how much risk for postoperative complications there was.
+Studies have investigated if ASA score also is a good predictor in other
+aspects of medical care and as discussed above it is a predictor for
+mortality after trauma and for readmission. However, this does not
+automatically mean that there is an correlation between ASA score and
+OFI. Patient's with higher ASA scores might receive more attention when
+admitted to the ICU due to their comorbidities, but as the variable used
+in this study is ASA preinjury --- a measurement of the patient's health
+before the trauma --- this might not contribute to a fair picture of the
+current state of the patient, which injury scores like RTS might.
 
@@
 treatment. Trying to reduce number of OFI can therefore not be solved by
-increasing medical resources outside working hours.
-
-#Lägga in artikel som pekar på om duty påverkar handläggning eller inte
+increasing medical resources outside working hours. On-call sleeping has
+been studied and it has been stated that sleeping between calls is
+beneficial for waking performance. However, there might be a temporary
+increase in performance impairment that should be taken into
+consideration [@35].
 
@@
 other parameters like vital signs, comorbidities and mechanism of
-injury.
+injury. ISS is an anatomical model, compared to RTS which is a
+physiological model. Among trauma patients admitted to the ICU the
+physiological models perform better than the anatomical models. However,
+patients with low scores in the physiological model also require an
+anatomic analysis of the injuries to determine their severity [@36].
+NISS has been found to be better than ISS in predicting the ICU
+admission of trauma patients, as well as for predicting prolonged ICU
+length of stay [@37]. This could be an explanation to why ISS is not
+significant in the results of this analysis, as RTS as well as NISS are
+potentially better models for patients admitted to the ICU.
 
-#Artikel om hur ISS bedöms kunna användas i samanhang av trauma
+### Strengths and limitations
 
-### Limitations
+One of the strengths of this study is that it is a registry-based study
+which means that the quality of the data is reliable. The collection of
+the included variables were pre-defined which means the reporting of the
+included variables was standardized. Furthermore, the specific inclusion
+criteria of the study: including patients 15 years and older, who were
+not dead on arrival and who were admitted to the ICU, resulted in a
+specific and well-defined cohort. This contributes to the reliability
+and reproducibility of the study.
 
-There were several limitations in our study: the selection for the
-morbidity and mortality review, the definition of the RTS variable and
-the study size.
+There were several limitations to our study: the limited study size, the
+selection of patients for the morbidity and mortality review and the
+definition and use of the physiological variables of the study.
 
-The selection for review for the morbidity and mortality review relied
-mostly on the local audit filters used at the Karolinska Institute
-Hospital in Solna. This leads to a risk of misplacement where some
-patients with OFI were not selected by for the review based on the audit
-filters although there was an OFI later on in the hospitalization
-process.
+Firstly, the study size was limited. This was a single-centre study and
+illustrate the situation at the Karolinska Intitute Hospital in Solna.
+Due to the small cohort of this study, it limits the applicability of
+this study to other level 1 trauma centers and ICUs. The small cohort
+might also affect the accuracy of the results.
 
-#Lägga in källa för audit filters - Kelvins artikel (mejlar Jocke)
+Secondly, the selection of patients for the morbidity and mortality
+review relied mostly on the local audit filters used at the Karolinska
+Institute Hospital in Solna. Using audit filters to select patients for
+the conference is associated with high false positive rates, ranging
+from 24% to 80% [@38]. Even though the use of audit filters is
+associated with high false positive rates, there is still a risk of
+misplacement where some patients with OFI were not selected by for the
+review based on the audit filters although there was an OFI that
+occurred later on during the hospital stay.
 
-The definition of the RTS-value is a combination of GCS, SBP and RR.
-When the patients were intubated and had missing values for GCS or RR,
-we replaced it with RTS with a value of 0 for the equation for RTS, as
-GCS and respiratory rate are 0 when patients are intubated. This could
-affect the final results, because then we would not get their initial
-values for these variables, meaning some patients received a RTS score
-more weighted towards SBP than they should have. However, the use of RTS
-can also be seen as an advantage, being a combination of vital
-parameters. This would mean that the RTS value reflects the state of the
-patient better than these parameters separately.
-
-The study was a single-centre study and illustrate the situation in the
-Karolinska Intitute Hospital in Solna. More research is needed on the
-area of OFI in patients admitted to the ICU. A suggestion would be to
-conduct a multi-centre study on this group of patients to further
-investigate areas for improvement and to further develop trauma quality
-programs. Based on the results from this study, variables of interest
-could be number of days in the ICU and investigation of GCS, RR and SBP
-separately instead of them as a combined RTS-score. A study including
-more patients could potentially show significance for the separate vital
-parameters leading to a better prediction model for which patients that
-should receive more focus when improving trauma care.
+The definition of the RTS-value is a combination of GCS, SBP and RR. The
+variables used were the values from the emergency department. When these
+were missing, the prehospital values were used. If these were also
+missing, but there was documentation indicating that the patients had
+been intubated, we replaced GCS and RR with a value of 0. This is
+motivated for due to patients not breathing or being conscious when
+intubated. The acquired values were then combined to calculate a value
+for RTS. If the patient had missing values for either one of GCS, SBP or
+RR they were excluded from the study. The handling of these variables
+and their combination to RTS could affect the result as the values used
+are from either prehospital values, emergency department values or from
+identification of intubation. This could potentially affect the accuracy
+of the data and decrease the possibility for reproducibility. However,
+one could also argue that as the RTS value is a combined score of 3
+physiological variables the combination of them decreases the overall
+inaccuracy and produces a satisfactory measurement of the physiological
+state of the patient.
 
@@
 absence of OFI. The external validity is therefore compromised due to
-the limited size of the Swedish trauma database of patients admitted to
-the ICU.
+the limited size of the study. The internal validity is however high
+thanks to the well-defined cohort specifying trauma patients admitted to
+the ICU at the Karolinska University Hospital in Solna. The results from
+this study is therefore highly relevant for understanding where further
+efforts for trauma improvement should be directed among ICU trauma
+patients at this hospital.
+
+### Implications
+
+The potential clinical implication for this study is to further
+contribute to the development of improved trauma care. Identifying the
+group of patients admitted to the ICU is another step in the direction
+of identifying the patient's who would benefit the most from further
+improvements in trauma care. Trauma has a great global impact and this
+study amongst others on the subject all contribute to the more specified
+efforts in trauma quality improvement systems. Additionally, this study
+has offered further insights to the patient profile of the trauma center
+at Karolinska University Hospital in Solna admitted to the ICU.
+
+### Future studies
+
+More research is needed on the area of OFI in patients admitted to the
+ICU. The most relevant limitation of this study was the limited study
+size. A suggestion would be to conduct a multi-centre study, covering
+larger databases or combining several databases, on this group of
+patients to further investigate areas for improvement and to further
+contribute to the improvement of trauma quality programs.
+
+Based on the results from this study, variables of interest for future
+studies could be number of days in the ICU and investigation of GCS, RR
+and SBP separately instead of them as a combined RTS-score. It would be
+interesting to see if one of the vital parameters is better for
+predicting OFI than the others in a larger cohort.
 
 Based on the statistical significance for higher RTS being associated
-with higher odds of OFI, there should be more focus on trauma patients
-who are stable upon arrival who have a risk of becoming worse. Future
-studies should investigate what factors to use to see the indication for
-such a focus. Focusing on patients who are stable upon arrival overall
-is very extensive. Therefore, identifying such factors would be
-valuable.
+with higher odds of OFI, efforts should be more focused towards trauma
+patients who are relatively stable upon arrival, but who have a risk of
+getting worse. Future studies should investigate what factors that could
+potentially indicate eventual worsening of the state of the patient.
+Focusing on patients who are stable upon arrival overall is very
+extensive, which further emphasizes the importance of identifying such
+variables.
+
+ASA preinjury did not show any correlation with OFI in this study.
+However, it would be interesting to further investigate the correlation
+between ASA preinjury or another scoring system of the patient's general
+health before admittance to the hospital, and a combined physiological
+and anatomical injury scoring system to see if they predict OFI better
+combined than separately.
 
@@
 
+# Contributions
+
+This study has been conducted by Elin Sun Cao with continuous guidance
+and feedback provided by Martin Gerdin Wärnberg, Jonatan Attergrim and
+Kelvin Szolnoky.
+
+# Acknowledgements
+
+I want to extend my greatest gratitude to my supervisor Martin Gerdin
+Wärnberg and co-supervisors Jonatan Attergrim and Kelvin Szolnoky who
+have provided me with guidance and support throughout this semester.
+Their extensive knowledge, deep insight into the subject and
+availability and support has greatly contributed to the development of
+this project. Additionally, I want to express my gratitude to my friends
+and family who have been there contributing with encouragement and
+positive energy.
+
 # References

### 2024-03-25 functions/table3.R

@@ -1,0 +1,45 @@
+#TABLE 2: Adjusted and unadjusted logistic regression
+# Data Preparation
+ofi_alive <- subset(ofi, subset = (Survival == "Dead"))
+
+tablereg1 <- ofi_alive %>% 
+  select(Sex, Age, Intubation, RTS, ISS,TimeFCT, OnDuty, daysinICU, TimeFCT, 
+         ASApreinjury, OpportunityForImprovement1)
+
+# Unadjusted Table
+table3aalive <- tbl_uvregression(data = tablereg1,
+                            method = glm,
+                            y = OpportunityForImprovement1,
+                            method.args = list(family = binomial),
+                            exponentiate = TRUE, 
+                            label = list(
+                              RTS = "Revised Trauma Score",
+                              daysinICU = "Days in the ICU",
+                              TimeFInt = "Time to first intervention",
+                              ASApreinjury = "ASA preinjury",
+                              OnDuty = "On duty",
+                              TimeFCT = "Time to first CT"
+                            )) %>%
+  bold_labels() %>%
+  bold_p(t = 0.05) 
+
+# Adjusted Table
+#Creating linear regression 
+adjusted_table1 <- glm(OpportunityForImprovement1 ~ Sex + Age + Intubation + RTS +  ISS + OnDuty + daysinICU + TimeFCT + ASApreinjury, family = binomial, data = tablereg1) 
+
+table3balive <- tbl_regression(adjusted_table1,
+                          exponentiate = TRUE, 
+                          label = list(RTS = "Revised Trauma Score",
+                                       daysinICU = "Days in the ICU",
+                                       ASApreinjury = "ASA preinjury",
+                                       TimeFCT = "Time to first CT")) %>%
+  bold_labels() %>%
+  bold_p(t = 0.05)
+
+# Merging Tables
+table3alive_merge <- tbl_merge(tbls = list(table3aalive, table3balive),
+                          tab_spanner = c("**Unadjusted**", "**Adjusted**")) %>%
+  modify_caption("<div style='text-align: left; font-weight: bold; color: black'>Table 3. Unadjusted and adjusted logistic regression analyses of associations between patient level factors and opportunities for improvement in patients alive 30 days after hospitalization</div>")
+
+print(table3alive_merge)
+

### 2024-03-27 manuscript.Rmd
27/3

@@
 title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
-output: "html_document"
+output: "word_document"
 author: "Elin Sun Cao"

### 2024-03-27 functions/table3.R

@@ -1,6 +1,6 @@
 #TABLE 2: Adjusted and unadjusted logistic regression
 # Data Preparation
-ofi_alive <- subset(ofi, subset = (Survival == "Dead"))
+ofi_alive <- subset(ofi, subset = (Survival == "Alive"))
 
 tablereg1 <- ofi_alive %>% 
   select(Sex, Age, Intubation, RTS, ISS,TimeFCT, OnDuty, daysinICU, TimeFCT, 
